---
title: "Data Preparation for HTNA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation for HTNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

This vignette shows how to go from raw event logs to a heterogeneous
transition network using `Nestimate::prepare()` for data preparation and
`htna::build_htna()` for network construction.

## Starting point: a raw event log

In practice, interaction data often arrives as a single table with one row per
event. Each row records which actor performed the action, what the action was,
and when it occurred.

We simulate a Human--AI tutoring interaction where each actor has four distinct
codes:

```{r}
library(htna)
library(Nestimate)

set.seed(123)

human_codes <- c("Ask", "Confirm", "Revise", "Evaluate")
ai_codes    <- c("Suggest", "Explain", "Execute", "Clarify")
sessions    <- paste0("S", sprintf("%03d", 1:30))

make_events <- function(codes, sessions, parity) {
  rows <- lapply(sessions, function(sid) {
    n <- sample(15:25, 1)
    data.frame(
      session_id       = sid,
      code             = sample(codes, n, replace = TRUE),
      order_in_session = seq_len(n) * 2L - parity,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

human_data <- make_events(human_codes, sessions, 1L)
ai_data    <- make_events(ai_codes, sessions, 0L)
```

We can also combine them into a single event log with an actor column, as
data often arrives in this form:

```{r}
events <- rbind(
  data.frame(actor = "Human", human_data, stringsAsFactors = FALSE),
  data.frame(actor = "AI", ai_data, stringsAsFactors = FALSE)
) 
events$actor = factor(events$actor, levels = c("Human","AI"))
events <- events[order(events$session_id, events$order_in_session), ]
head(events[, c("actor", "session_id", "code", "order_in_session")])
```

## Option 1: Single data frame with an actor column

If your data is already in long format with session and order columns,
you can pass it directly to `build_htna()` using the `actor_col` argument:

```{r}
net <- build_htna(events, actor_col = "actor")
net
```

```{r}
plot_htna(net)
```

## Option 2: Named list of data frames

When data for each actor is stored separately (or has been prepared
independently), pass them as a named list:

```{r}
net2 <- build_htna(list(Human = human_data, AI = ai_data))
net2
```

## Option 3: Prepare each actor with `Nestimate::prepare()`

When your raw data needs cleaning -- e.g. session detection from timestamps,
tie-breaking, or format conversion -- you can use `Nestimate::prepare()` on
each actor's data independently, then feed the prepared long-format data
into `build_htna()`.

```{r}
prep_human <- prepare(
  human_data,
  actor   = "session_id",
  action  = "code",
  session = "session_id",
  order   = "order_in_session"
)

prep_ai <- prepare(
  ai_data,
  actor   = "session_id",
  action  = "code",
  session = "session_id",
  order   = "order_in_session"
)
```

`prepare()` returns a list with several components. The `long_data` element
contains the cleaned long-format data ready for network estimation:

```{r}
head(prep_human$long_data[, c("session_id", "code", "order_in_session")])
```

Pass the prepared long data as a named list:

```{r}
net3 <- build_htna(
  list(Human = prep_human$long_data, AI = prep_ai$long_data)
)
net3
```

```{r}
plot_htna(net3, layout = "auto")
```

## When to use each approach

| Approach | Best for |
|----------|----------|
| Single data frame + `actor_col` | Data already clean, sessions and order defined |
| Named list of data frames | Separate files per actor, no cleaning needed |
| Separate `prepare()` per actor | Raw timestamps, automatic session splitting, different cleaning per actor |

All three approaches produce the same network when the underlying data is
equivalent. Choose whichever fits your data pipeline.

## Handling overlapping codes

If different actors share the same code labels (e.g. both Human and AI have a
code called "Ask"), `build_htna()` will raise an error by default. Set
`disambiguate = TRUE` to automatically prefix codes with the actor name:

```{r, eval = FALSE}
net <- build_htna(data, actor_col = "actor", disambiguate = TRUE)
# Codes become "Human:Ask", "AI:Ask", etc.
```
