---
title: "State frequencies and chi-square mosaics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State frequencies and chi-square mosaics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 8, fig.height = 5, dpi = 100
)
```

# Background

Two complementary classes of summary are commonly applied to a
transition network: descriptive summaries of the marginal state
distribution, and inferential comparisons of the joint transition
distribution against an independence baseline. The first class
characterises which states occur and in what proportion; the second
identifies transitions that are over- or under-represented relative
to what independence between source and target would predict.

These classes belong to different statistical regimes and therefore
warrant different visualisations. Marginal summaries (treemaps,
bar charts, frequency tables) are descriptive and require only the
empirical state counts. Joint summaries (chi-square mosaics) are
inferential and require integer transition counts in order to
compute residuals against an independence model.

This vignette presents the htna interfaces to both classes:

- The frequency family — `frequencies_htna()`,
  `plot_frequencies_htna()`, `state_frequencies_htna()` — for
  marginal state-level summaries.
- The mosaic family — `mosaic_plot_htna()` — for chi-square
  comparisons of the joint transition matrix against independence.

# Data and networks

Two networks are constructed from the `Human` / `AI` example
sequences. The frequency family operates on either, but the mosaic
family requires an integer-weighted network because the chi-square
test acts on counts; this is the network produced under
`method = "frequency"`.

```{r load}
library(htna)
data(human_long, ai_long, package = "Nestimate")

net      <- build_htna(list(Human = human_long, AI = ai_long))
net_freq <- build_htna(list(Human = human_long, AI = ai_long),
                       method = "frequency")
```

# Marginal state distributions

## Tabular summary

`frequencies_htna()` returns the per-actor marginal state
distribution as a data frame with one row per (actor, state) pair.
Columns are: actor (`group`), state, count, and within-network
proportion. The table is the data underlying every chart in this
section.

```{r}
frequencies_htna(net)
```

## Graphical summary: `plot_frequencies_htna()`

`plot_frequencies_htna()` renders the same marginal distribution in
three layouts. Each layout encodes the same data but is optimised
for a different reading task.

### Treemap (`view = "treemap"`, default)

Each panel corresponds to one actor; tile area within a panel
encodes the within-network proportion of the corresponding state.
The layout is space-efficient and allows the full state vocabulary
to be displayed simultaneously.

```{r}
plot_frequencies_htna(net)
```

The function returns a `state_freq` object (a list containing the
rendered chart and the underlying frequency table) invisibly. Any
argument accepted by `Nestimate::plot_state_frequencies()` may be
passed through `...`.

### Combined bars (`view = "bars"`)

The bars layout collapses both actors onto a single y-axis, sorts
the states by total count, and colours the bars by actor. This
layout is appropriate when the analytic task is direct numerical
comparison across the full vocabulary.

```{r}
plot_frequencies_htna(net, view = "bars")
```

The bars layout returns a ggplot object, permitting modification
through standard ggplot composition operators:

```{r}
plot_frequencies_htna(net, view = "bars") +
  ggplot2::labs(title = "Human vs AI: state counts")
```

### Per-actor faceted bars (`view = "facet"`)

The faceted layout assigns each actor its own panel with an
independent y-axis. This is appropriate when the actors differ
substantially in event volume, since a shared scale would compress
the lower-volume actor's bars beyond legibility.

```{r}
plot_frequencies_htna(net, view = "facet")
```

Like `view = "bars"`, this returns a ggplot object.

## Raw-data alias: `state_frequencies_htna()`

When the input is a vector of state codes rather than a constructed
network, `state_frequencies_htna()` returns the same `state /
count / proportion` summary without requiring `build_htna()` to be
called first. This is useful for preliminary data inspection prior
to network construction.

```{r}
state_frequencies_htna(human_long$code)
```

# Joint transition distribution: chi-square mosaic

`mosaic_plot_htna()` displays the joint distribution of `(source,
target)` transitions as a chi-square mosaic. Each cell of the
transition matrix is rendered as a rectangle whose area is
proportional to the joint share of that transition; cell colour
encodes the standardised residual against an independence model.
Cells with positive residuals (over-represented relative to
independence) are coloured blue; cells with negative residuals
(under-represented) are coloured red; cells whose observed value
matches the independence prediction are white.

The chi-square test requires integer counts, hence the requirement
for a frequency-method network.

## Default residuals (permutation-based)

The default residual estimator is permutation-based, with
`n_perm = 500` iterations. Permutation residuals are appropriate
when cell counts are sparse or when the chi-square asymptotic
approximation is not trusted; the trade-off is computation time.

```{r}
mosaic_plot_htna(net_freq, seed = 1L)
```

Cells with strong positive residuals identify transitions that
characterise the process beyond what would be expected by chance;
cells with strong negative residuals identify transitions that are
systematically suppressed.

## Asymptotic residuals (`residuals = "asymptotic"`)

The closed-form chi-square standardised residual estimator
(`chisq.test()$stdres`) is faster and is the convention used by
`tna` and `vcd`. It is appropriate when cell counts are large
enough for the asymptotic approximation to hold.

```{r}
mosaic_plot_htna(net_freq, residuals = "asymptotic")
```

For htna corpora of typical size (hundreds of sessions, hundreds
to thousands of transitions per cell), permutation and asymptotic
residuals agree closely.

## Colour-scale clipping (`range = c(-4, 4)`)

By default the colour scale is calibrated to the maximum absolute
residual in the matrix. A single extreme cell can therefore
desaturate the remainder of the chart. Clipping the range to a
fixed interval preserves contrast across the matrix and supports
comparison of mosaics across networks.

```{r}
mosaic_plot_htna(net_freq, range = c(-4, 4), seed = 1L)
```

A range of ±4 is conventional in mosaic displays; residuals beyond
the range saturate to the most intense colour.

## Axis-label rotation (`top_angle`, `left_angle`)

For matrices with long state names or large vocabularies, axis
labels may collide. The `top_angle` and `left_angle` arguments
rotate the labels for legibility.

```{r}
mosaic_plot_htna(net_freq, top_angle = 45, left_angle = 0, seed = 1L)
```

These are layout adjustments and do not affect the underlying
statistical computation.

# Selection criteria

| Analytic question | Recommended chart |
| --- | --- |
| Per-actor state mix as proportions | `plot_frequencies_htna(net)` (treemap) |
| Cross-corpus rank ordering of codes | `plot_frequencies_htna(net, view = "bars")` |
| Per-actor mix when event volumes differ substantially | `plot_frequencies_htna(net, view = "facet")` |
| Numerical access to the marginal distribution | `frequencies_htna(net)` |
| Identification of over- and under-represented transitions | `mosaic_plot_htna(net_freq, seed = 1L)` |
| Comparison of mosaics across networks | `mosaic_plot_htna(net_freq, range = c(-4, 4))` |

# Related vignettes

The companion `sequence-plots` vignette covers time-preserving
displays of the same data (per-session trajectories). The
`model-assessment` vignette covers reliability and stability
analyses of the network estimate.
