---
title: "Grouped HTNA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped HTNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
```

A common HTNA question is whether the same actor partition behaves
*differently* across experimental conditions -- e.g. a control group
versus a treatment group. `build_htna()` supports this directly: pass
the name of a session-level grouping column and you get **one network
per condition**, each preserving the actor partition. Every other
function in `htna` recognises the grouped result and iterates over
conditions.

## A control / experimental example

We simulate a study with two conditions, each with 30 sessions. In
every session a Human and an AI interleave codes from the **same**
alphabet, but the experimental condition uses a more collaborative
transition pattern (more `Specify -> Plan` instead of `Specify ->
Execute`). Because both conditions share the same nodes, we can compare
them edge by edge.

```{r}
library(htna)
set.seed(42)

human_codes <- c("Specify", "Verify", "Refine")
ai_codes    <- c("Plan", "Execute", "Explain")

# Markov-style transition matrices per condition.
# Rows = previous state, columns = next state. Both conditions share
# the same alphabet, but Control favours fast Specify -> Execute loops
# while Experimental favours deliberative Specify -> Plan -> Verify.
states_ <- c("Specify", "Verify", "Refine", "Plan", "Execute", "Explain")

trans_control <- matrix(c(
  # Specify  Verify Refine  Plan Execute Explain
    0.05,    0.05,  0.05,   0.10, 0.70,   0.05,   # from Specify
    0.10,    0.05,  0.30,   0.05, 0.45,   0.05,   # from Verify
    0.10,    0.10,  0.10,   0.05, 0.60,   0.05,   # from Refine
    0.05,    0.10,  0.10,   0.05, 0.65,   0.05,   # from Plan
    0.40,    0.20,  0.30,   0.05, 0.05,   0.00,   # from Execute
    0.05,    0.50,  0.05,   0.10, 0.25,   0.05    # from Explain
), nrow = 6, byrow = TRUE,
dimnames = list(states_, states_))

trans_experimental <- matrix(c(
  # Specify  Verify Refine  Plan Execute Explain
    0.05,    0.10,  0.05,   0.65, 0.05,   0.10,   # from Specify
    0.05,    0.05,  0.55,   0.10, 0.05,   0.20,   # from Verify
    0.10,    0.30,  0.05,   0.40, 0.10,   0.05,   # from Refine
    0.05,    0.50,  0.05,   0.05, 0.05,   0.30,   # from Plan
    0.05,    0.50,  0.05,   0.30, 0.05,   0.05,   # from Execute
    0.05,    0.30,  0.10,   0.45, 0.05,   0.05    # from Explain
), nrow = 6, byrow = TRUE,
dimnames = list(states_, states_))

simulate_session <- function(sid, condition, alphabet, trans) {
  n_steps <- sample(8:24, 1)            # variable session length
  states  <- character(n_steps)
  states[1] <- sample(alphabet, 1)
  for (i in seq_len(n_steps - 1L)) {
    states[i + 1L] <- sample(alphabet, 1, prob = trans[states[i], ])
  }
  human_set <- human_codes
  data.frame(
    session_id = sid,
    code       = states,
    actor      = ifelse(states %in% human_set, "Human", "AI"),
    order_in_session = seq_len(n_steps),
    condition  = condition,
    stringsAsFactors = FALSE
  )
}

alphabet <- c(human_codes, ai_codes)

ctrl_data <- do.call(rbind, lapply(
  paste0("C", sprintf("%03d", 1:30)),
  simulate_session, condition = "Control",
  alphabet = alphabet, trans = trans_control))

exp_data  <- do.call(rbind, lapply(
  paste0("E", sprintf("%03d", 1:30)),
  simulate_session, condition = "Experimental",
  alphabet = alphabet, trans = trans_experimental))

events <- rbind(ctrl_data, exp_data)
head(events)
```

## Building one network per condition

`build_htna()` takes the long data frame, the node-to-actor lookup
(`node_groups`), and a `group` argument naming the session-level
condition column:

```{r}
grp <- build_htna(
  events,
  node_groups = list(Human = human_codes, AI = ai_codes),
  group       = "condition"
)
length(grp)
names(grp)
```

The result is a named list of HTNA networks. Indexing one element gives
a regular HTNA network you can use exactly like any other:

```{r}
grp$Control
```

Because both conditions were built from the same alphabet, both
networks have the **same nodes** -- which is exactly what
`plot_htna_diff()` and `permutation()` need.

## Plotting per condition

`plot_htna()` iterates over conditions and draws one network per
element. It does not manage the layout grid -- wrap with
`par(mfrow = ...)` if you want a panel:

```{r, fig.height = 5}
op <- par(mfrow = c(1, 2))
plot_htna(grp)
par(op)
```

## Per-actor sequences per condition

`sequence_plot_htna()` works the same way:

```{r, fig.height = 8}
sequence_plot_htna(grp, by = "group", type = "index")
```

## Centralities per condition

`centralities(grp)` returns one tidy data frame, with a leading
`group` column identifying the condition:

```{r}
ct <- centralities(grp)
head(ct, 10)
```

`plot_centralities()` faces a grid: rows are conditions, columns are
measures, so condition-level differences are easy to read at a glance:

```{r, fig.height = 9}
plot_centralities(grp, by = "group")
```

## Bootstrap per condition

`bootstrap_htna(grp)` runs the bootstrap on each condition and returns
a group of bootstrap results. `plot_htna_bootstrap()` iterates again:

```{r, fig.height = 5}
boots <- bootstrap_htna(grp, iter = 200)
op <- par(mfrow = c(1, 2))
plot_htna_bootstrap(boots, display = "significant")
par(op)
```

## Comparing the two conditions

This is the key payoff of a control / experimental design: both
networks share an alphabet, so you can compute the elementwise
difference directly with `plot_htna_diff()`. Positive (experimental >
control) is green, negative (control > experimental) is red:

```{r}
plot_htna_diff(grp$Experimental, grp$Control)
```

`permutation()` quantifies which differences are statistically
significant:

```{r}
perm <- permutation(grp$Experimental, grp$Control, iter = 500)
plot_htna_diff(perm)
```

```{r}
plot_htna_diff(perm, show_nonsig = TRUE)
```

The significant edges that pop out are exactly the transitions the
simulation was designed to bias -- in particular the
`Specify -> Plan` versus `Specify -> Execute` switch -- which
demonstrates the workflow on a controlled example.

## Indexing

Because the result of `build_htna(..., group = ...)` is a named list,
all standard list operations work: `grp[["Control"]]`,
`grp[c("Control", "Experimental")]`, `length(grp)`, `names(grp)`.
Subsets keep the grouped class so they pass straight back to the
wrappers above.
