---
title: "Getting Started with htna"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with htna}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
```

Heterogeneous Transition Network Analysis (HTNA) studies sequences in
which two or more actors interleave -- a learner and a tutor, a human
and an AI, a clinician and a patient -- and treats each actor's codes as
a distinct node group. The `htna` package builds the network, computes
the usual analytical quantities, and renders the result with the actor
partition baked into colour, layout, and the legend.

## 1. Building a heterogeneous network

Start from one long-format data frame with an actor-type column tagging
each row (`"Human"` or `"AI"`) and pass its name as `actor_type` to
`build_htna()`. The result is a network whose nodes carry the actor
label they came from. The bundled `human_ai` corpus (see `?human_ai`)
is used throughout this vignette.

```{r}
library(htna)
data(human_ai)

net <- build_htna(human_ai, actor_type = "actor_type")
```

## 2. Plotting the network

`plot_htna()` lays out actor groups around the circle, colours each node
by its actor type, and draws the actor legend below the plot:

```{r}
plot_htna(net)
```

## 3. Per-actor sequence plots

`sequence_plot_htna()` shows the temporal structure of the sessions.
With `by = "state"` (the default) each cell is coloured by its 
code; with `by = "group"` cells are coloured by actor type. `type`
selects the layout: `"index"` repreents the index plot, `"heatmap"`
collapses across sessions into a single carpet, and `"distribution"`
shows the sequence distribution plot with the state composition per timepoint 
as a stacked area.

```{r, fig.height = 5}
sequence_plot_htna(net, by = "state", type = "index")
```

```{r}
sequence_plot_htna(net, by = "state", type = "heatmap")
```

```{r, fig.height=4}
sequence_plot_htna(net, by = "group", type = "distribution", na_color = "white")
```

When `by = "state"`, the legend is split into one block per actor type
with the actor type name above each block, so the reader can tell at
a glance which codes belong to which actor type. 

## 4. Centralities

`centralities_htna()` calculates per-node centrality measure: one row per node, 
one column per measure, defaulting to nine standard measures (out/in strength,
in/out closeness, closeness, betweenness, RSP betweenness, diffusion,
clustering). The `measures = c(...)` argument can be used to specify a different 
set of measures

```{r}
centralities_htna(net)
```

To plot centralities, the `plot_centralities()` function can be used. 
Each panel is one measure; bars are coloured per state by default, 
or by actor type with `by = "group"`:

```{r, fig.height = 9}
plot_centralities(net, by = "state")
```

```{r, fig.height = 9}
plot_centralities(net, by = "group")
```

## 5. Bootstrap

To validate which transitions of an HTNA model are stable, `bootstrap_htna()` 
resamples sessions to  obtain edge-weight stability and per-edge p-values. 
The `plot_htna_bootstrap()` plots the resampled network with confidence intervals and a non-significant/significant
edge style. The argument `display = "significant"` shows also non-significant edges.

```{r}
boot <- bootstrap_htna(net)
plot_htna_bootstrap(boot)
```

```{r}
plot_htna_bootstrap(boot, display = "significant")
```

## 6. Centrality stability

The function `centrality_stability_htna()` runs a case-dropping centrality
stability check: drops a proportion
of the sessions, recomputes the centralities, and measures how strongly
the dropped-sample centralities correlate with the originals. The
returned `cs` value per measure is the **largest drop proportion at
which the correlation with the original centralities still meets the
threshold** (default `0.7`) for at least `certainty` (default `0.95`)
of resamples — values above ~0.5 are typically considered stable.

```{r}
cs <- centrality_stability_htna(net, iter = 100, seed = 1)
cs$cs
```

By default the check covers `InStrength`, `OutStrength`, and
`Betweenness` — the three measures whose values are bit-equal between
htna's `cograph` engine and the reference implementation. 
 

## 7. Split-half reliability

The function `reliability_htna()` reports how stable each edge weight is under
random split-half resampling: it draws `iter` random splits of the
sessions, builds a network on each half, and summarises the
agreement (mean / median / max absolute deviation, plus correlation)
between paired half-networks.  

```{r}
rel <- reliability_htna(net, iter = 100, seed = 1)
rel$summary
```


## 8. Comparing two networks

Given two HTNA networks built over the same node set, `plot_htna_diff()`
draws the edge-level pairwise difference. Positive differences are green,
negative red.

For example, we compare early vs. late sessions:

```{r}
sess_start <- aggregate(session_date ~ session_id, data = human_ai,
                        FUN = min)
sess_start <- sess_start[order(sess_start$session_date,
                               sess_start$session_id), ]
half       <- nrow(sess_start) %/% 2L
early_sess <- sess_start$session_id[seq_len(half)]

early <- build_htna(human_ai[human_ai$session_id %in% early_sess, ],
                    actor_type = "actor_type")
late  <- build_htna(human_ai[!human_ai$session_id %in% early_sess, ],
                    actor_type = "actor_type")

plot_htna_diff(early, late)
```

## 9. Permutation test

The function `permutation_htna()` provides a non-parametric significance test on each
edge weight difference. Pass the result to `plot_htna_diff()` to render
significant differences with the pos/neg colouring above; non-significant
edges are dashed grey when `show_nonsig = TRUE`.

```{r}
perm <- permutation_htna(early, late, iter = 200)
plot_htna_diff(perm)
```

```{r}
plot_htna_diff(perm, show_nonsig = TRUE)
```

## 10. Path patterns

`extract_meta_paths()` discovers recurring patterns at two levels. By
default it returns concrete state-level patterns (alphabet = the code
set `Ask`, `Plan`, `Check`, ...) and annotates each row with the
type-level template it instantiates:

```{r}
extract_meta_paths(net, length = 3)
```

Pass `level = "type"` for the type-level meta-path summary - a fixed
sequence of actor labels that abstracts away from concrete states
(e.g. `Human->AI->Human`):

```{r}
extract_meta_paths(net, level = "type", length = 3)
```

A `schema` filters the search. Each part can be a type name (expands to
every code in that group), a concrete state, or `*`:

```{r}
extract_meta_paths(net, schema = "Human->AI->Human")
extract_meta_paths(net, schema = "Human->Ask->*")
```

Filter by lift to surface over-represented patterns (lift > 1 means
more frequent than independence would predict):

```{r}
extract_meta_paths(net, length = 3, min_lift = 2)
extract_meta_paths(net, level = "type", length = 3, min_lift = 1.2)
```
