---
title: "Getting Started with htna"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with htna}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
```

Heterogeneous Transition Network Analysis (HTNA) studies sequences in
which two or more actors interleave -- a learner and a tutor, a human
and an AI, a clinician and a patient -- and treats each actor's codes as
a distinct node group. The `htna` package builds the network, computes
the usual analytical quantities, and renders the result with the actor
partition baked into colour, layout, and the legend.

## 1. Building a heterogeneous network

Start from one long-format data frame per actor and combine them with
`build_htna()`. The result is a network whose nodes carry the actor
label they came from:

```{r}
library(htna)

data(human_long, ai_long, package = "Nestimate")

net <- build_htna(list(Human = human_long, AI = ai_long))
net
```

## 2. Plotting the network

`plot_htna()` lays out actor groups around the circle, colours each node
by its actor, and draws the actor legend below the plot:

```{r}
plot_htna(net)
```

## 3. Per-actor sequence plots

`sequence_plot_htna()` shows the temporal structure of the sessions.
With `by = "state"` (the default) each cell is coloured by its concrete
code; with `by = "group"` cells are coloured by actor.

```{r, fig.height = 5}
sequence_plot_htna(net, by = "state", type = "index")
```

```{r}
sequence_plot_htna(net, by = "state", type = "heatmap")
```

```{r}
sequence_plot_htna(net, by = "group", type = "distribution")
```

When `by = "state"`, the legend is split into one block per actor with
the actor name above each block, so the reader can tell at a glance
which codes belong to which actor.

## 4. Centralities

`centralities()` returns a tidy data frame: one row per node, one column
per measure, defaulting to nine standard measures (out/in strength,
in/out closeness, closeness, betweenness, RSP betweenness, diffusion,
clustering). Pass `measures = c(...)` to subset.

```{r}
centralities(net)
```

`plot_centralities()` is the corresponding plot. Each panel is one
measure; bars are coloured per state by default, or by actor with
`by = "group"`:

```{r, fig.height = 9}
plot_centralities(net, by = "state")
```

```{r, fig.height = 9}
plot_centralities(net, by = "group")
```

## 5. Bootstrap

`bootstrap_htna()` resamples sessions to obtain edge-weight stability
and per-edge p-values. `plot_htna_bootstrap()` plots the resampled
network with confidence intervals and a non-significant/significant
edge style. `display = "significant"` zeros out non-significant edges.

```{r}
boot <- bootstrap_htna(net)
plot_htna_bootstrap(boot)
```

```{r}
plot_htna_bootstrap(boot, display = "significant")
```

## 6. Comparing two networks

Given two HTNA networks built over the same node set, `plot_htna_diff()`
draws the elementwise difference. Positive differences are green,
negative red.

```{r}
n_h    <- nrow(human_long)
n_a    <- nrow(ai_long)
early  <- build_htna(list(Human = human_long[1:(n_h %/% 2), ],
                          AI    = ai_long[1:(n_a %/% 2), ]))
late   <- build_htna(list(Human = human_long[(n_h %/% 2 + 1):n_h, ],
                          AI    = ai_long[(n_a %/% 2 + 1):n_a, ]))

plot_htna_diff(early, late)
```

## 7. Permutation test

`permutation()` provides a non-parametric significance test on each
edge weight difference. Pass the result to `plot_htna_diff()` to render
significant differences with the pos/neg colouring above; non-significant
edges are dashed grey when `show_nonsig = TRUE`.

```{r}
perm <- permutation(early, late, iter = 200)
plot_htna_diff(perm)
```

```{r}
plot_htna_diff(perm, show_nonsig = TRUE)
```

## 8. Meta-paths

Meta-paths are type-level patterns: a fixed sequence of actor labels
that abstracts away from the concrete states. For example
`Human->AI->Human` is every length-3 path starting and ending in Human
states with an AI state in between.

```{r}
extract_meta_paths(net, length = 3)
```

Schemas accept a `*` wildcard to fix some positions while letting others
vary:

```{r}
extract_meta_paths(net, schema = "Human->*->Human")
```

Filter by lift to surface over-represented templates (lift > 1 means
more frequent than independence would predict):

```{r}
extract_meta_paths(net, length = 3, min_lift = 1.2)
```

## 9. State-level paths

`extract_paths()` is the state-level counterpart: same arguments, but
the alphabet is the concrete code set (`Ask`, `Plan`, `Verify`, ...)
rather than the actor types. Use it to discover the most frequent
concrete sequences:

```{r}
extract_paths(net, length = 3)
```

Wildcard schemas pin down a start and end while letting the middle vary:

```{r}
extract_paths(net, schema = "Specify->*->Verify")
```

And the lift filter surfaces concrete sequences that occur more often
than the state marginals would predict:

```{r}
extract_paths(net, length = 3, min_lift = 2)
```
