---
title: "Getting Started with htna"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with htna}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

Heterogeneous Transition Network Analysis (HTNA) extends traditional TNA to
sequences that mix two or more actor groups (e.g. Human and AI) within the same
interaction. The `htna` package provides a thin, focused API on top of the
[Nestimate](https://CRAN.R-project.org/package=Nestimate) estimation engine and
the [cograph](https://CRAN.R-project.org/package=cograph) rendering engine.

## Building a heterogeneous network

Start by loading the package and the example data shipped with Nestimate:

```{r}
library(htna)
data(human_long, ai_long, package = "Nestimate")
```

`build_htna()` accepts a named list of long-format data frames, one per actor
group. It combines the sequences, estimates transition probabilities, and stores
the actor partition on the resulting network object:

```{r}
net <- build_htna(list(Human = human_long, AI = ai_long))
net
```

The actor partition is available in two forms:

- `net$nodes$groups` -- actor label per node (auto-detected by `plot_htna()`).
- `net$node_groups` -- canonical `(node, group)` data frame used by
  `cograph::get_groups()` and `cograph::cluster_summary()`.

## Plotting

`plot_htna()` auto-detects the actor groups and renders them with distinct
colours. It defaults to a circular layout and a 0.05 edge threshold:
```{r}
plot_htna(net)
```

## Meta-paths

Meta-paths are type-level patterns that abstract away from concrete states and
ask how often a typed template appears in the data. For example,
`Human->AI->Human` captures every length-3 path that starts with a Human code,
transitions to an AI code, and returns to a Human code.

```{r}
# Enumerate all meta-paths of length 2 to 4
mp <- extract_meta_paths(net)
mp
```

You can search for a specific schema (with wildcards):

```{r}
# Any length-3 path that returns to Human
extract_meta_paths(net, schema = "Human->*->Human")
```

Gapped meta-paths allow skipping positions:

```{r}
extract_meta_paths(net, length = 3, type = "gapped", gap = 1:2)
```

Filter by lift to find over-represented meta-paths:

```{r}
extract_meta_paths(net, length = 3, min_lift = 1.2)
```

