---
title: "Input formats for build_htna()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input formats for build_htna()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

`build_htna()` accepts the same Human / AI sequence data in **three
interchangeable shapes**: a named list of per-actor frames, a single
combined frame with an actor-type column, or a single combined frame
with a node-to-actor lookup. This vignette walks through each form
using Nestimate's `human_long` + `ai_long` example and verifies that
all three produce the same network, then covers two more parameters
that compose with any input form: `actor` (forwarded to
`Nestimate::build_network()`) and `disambiguate` (for code-label
collisions).

## Setup

```{r}
library(htna)
data(human_long, ai_long, package = "Nestimate")

# What the two long frames look like
str(human_long, max.level = 1, give.attr = FALSE)
str(ai_long,    max.level = 1, give.attr = FALSE)
```

Both frames share the canonical schema htna expects: `session_id`,
`code`, `order_in_session`. Within each session, `order_in_session`
is a **shared** time index across both actors — Human's events occupy
the odd positions, AI's the even ones — so concatenating and sorting
recovers the temporal interleaving.

## Form 1 — Named list of per-actor frames

The most direct form. Each entry is one actor's long frame; the list
name becomes the actor-type label on the network.

```{r}
net_list <- build_htna(list(Human = human_long, AI = ai_long))
summary(net_list)
```

```{r, fig.height = 7}
plot_htna(net_list)
```

Use this form when each actor's data lives in a separate frame (the
typical case when the two streams were prepared independently).

## Form 2 — Single combined frame with `actor_type`

When the data already arrives as one long table, tag each row with the
actor-type and pass the column name to `actor_type =`.

```{r}
combined <- rbind(
  cbind(actor_type = "Human", human_long[, c("session_id", "code",
                                              "order_in_session")]),
  cbind(actor_type = "AI",    ai_long[,    c("session_id", "code",
                                              "order_in_session")])
)
combined$actor_type <- factor(combined$actor_type, levels = c("Human", "AI"))
combined <- combined[order(combined$session_id, combined$order_in_session), ]

head(combined)

net_actor_type <- build_htna(combined, actor_type = "actor_type")
```

The actor-type order on the resulting network follows the factor levels
(or, for character vectors, the order of first appearance), which is
how the actor groups will be ordered in legends and plots.

## Form 3 — Single combined frame with `node_groups`

When the combined frame has **no** actor-type column, declare the
node-to-actor partition directly via `node_groups`. `node_groups` accepts
two interchangeable shapes:

### 3a. Named list of code vectors

```{r}
human_codes <- unique(human_long$code)
ai_codes    <- unique(ai_long$code)

net_node_groups <- build_htna(
  combined[, c("session_id", "code", "order_in_session")],
  node_groups = list(Human = human_codes, AI = ai_codes)
)
```

### 3b. Two-column data frame (codebook)

The same partition expressed as a tidy codebook — one row per code, one
column for the action and one for the actor type. The action column must
match the `action` parameter (default `"code"`); the other column is
treated as the actor-type tag. Column order is irrelevant.

```{r}
codebook <- data.frame(
  code       = c(human_codes, ai_codes),
  actor_type = c(rep("Human", length(human_codes)),
                 rep("AI",    length(ai_codes))),
  stringsAsFactors = FALSE
)
head(codebook)

net_node_groups_df <- build_htna(
  combined[, c("session_id", "code", "order_in_session")],
  node_groups = codebook
)
```

The actor-type ordering follows the column's factor levels (if a
factor) or the order of first appearance (if a character vector). Use
this form when your code/actor mapping already lives in a tidy lookup
table.

## All four forms produce the same network

`build_htna()` resolves every input shape to the same combined sequence
and forwards to `Nestimate::build_network()`. To make the equivalence
verifiable rather than hand-waved, we compare the **wide session
matrix** (`$data`, the input that every downstream analysis consumes)
plus all the analytical slots — cell-for-cell, against the named-list
reference.

```{r}
forms <- list(
  list_form        = net_list,
  actor_type_form  = net_actor_type,
  node_groups_list = net_node_groups,
  node_groups_df   = net_node_groups_df
)

ref <- forms$list_form

# Helper: compare any slot from `other` to its counterpart in `ref`
# after aligning by names where applicable. Returns TRUE iff bit-equal.
slot_equal <- function(ref_v, other_v) {
  if (is.null(ref_v) && is.null(other_v)) return(TRUE)
  if (is.null(ref_v) ||  is.null(other_v)) return(FALSE)
  if (is.data.frame(ref_v)) {
    key <- intersect(c("node", "label"), names(ref_v))[1L]
    if (!is.na(key))
      other_v <- other_v[match(ref_v[[key]], other_v[[key]]), , drop = FALSE]
    rownames(other_v) <- rownames(ref_v) <- NULL
    return(identical(other_v, ref_v))
  }
  if (!is.null(dim(ref_v))) {
    if (!is.null(rownames(ref_v)) && !is.null(rownames(other_v)))
      other_v <- other_v[rownames(ref_v), colnames(ref_v), drop = FALSE]
    return(isTRUE(all.equal(unname(other_v), unname(ref_v))))
  }
  if (!is.null(names(ref_v)) && !is.null(names(other_v)))
    other_v <- other_v[names(ref_v)]
  isTRUE(all.equal(unname(other_v), unname(ref_v)))
}

# Cell-by-cell `$data` comparison: same shape, same row/col order,
# same value (or NA) at every position.
data_identical <- function(ref_d, other_d) {
  if (!identical(dim(ref_d), dim(other_d)))           return(FALSE)
  if (!identical(rownames(ref_d), rownames(other_d))) return(FALSE)
  if (!identical(colnames(ref_d), colnames(other_d))) return(FALSE)
  m1 <- as.matrix(ref_d);   dimnames(m1) <- NULL
  m2 <- as.matrix(other_d); dimnames(m2) <- NULL
  identical(m1, m2)
}

# Pairwise check vs the named-list reference
do.call(rbind, lapply(names(forms)[-1L], function(nm) {
  net <- forms[[nm]]
  data.frame(
    form         = nm,
    data_eq      = data_identical(ref$data, net$data),
    weights_eq   = slot_equal(ref$weights,     net$weights),
    initial_eq   = slot_equal(ref$initial,     net$initial),
    nodes_eq     = slot_equal(ref$nodes,       net$nodes),
    node_grps_eq = slot_equal(ref$node_groups, net$node_groups),
    actor_lvls_eq = identical(ref$actor_levels, net$actor_levels),
    stringsAsFactors = FALSE
  )
}))
```

Every cell reads `TRUE` — including the wide session matrix `$data`,
which is the single object every downstream function (`tna()`,
sequence plots, bootstrap, permutation, edge betweenness) actually
consumes. The input shape only affects how you express the data;
the produced network is bit-identical.

The same checks are run as automated tests in
`tests/testthat/test-tna-equivalence.R` across nine diverse datasets
(small/large alphabets, imbalanced actors, disconnected components,
variable session lengths, near-deterministic transitions, and the
real-world `tna::group_regulation_long`).

## Adding an individual-actor column with `actor =`

`actor_type` partitions **codes** into actor *groups* (Human vs AI).
The independent `actor =` parameter is forwarded to
`Nestimate::build_network()` with the same role it has there:
the column name identifying the **individual actor** that performed
each event (e.g. a learner / user id). It composes with any of the
three input forms above.

```{r}
# Treat session_id as the individual-actor key (one per conversation)
combined_with_actor <- combined
combined_with_actor$user_id <- combined_with_actor$session_id

net_with_actor <- build_htna(
  combined_with_actor,
  actor_type = "actor_type",
  actor      = "user_id"
)
net_with_actor
```

When `actor =` is omitted, Nestimate aggregates events at the session
level (the default htna behaviour). When it's supplied, Nestimate
groups events by individual actor first.

## Handling code-label collisions with `disambiguate =`

If both actor types share a code label (e.g. both Human and AI have a
code called `Ask`), `build_htna()` errors by default — keeping such
states distinct is usually what you want. Pass `disambiguate = TRUE`
to prefix every code with its actor-type label so the two `Ask` nodes
become `Human:Ask` and `AI:Ask`.

```{r}
overlap <- data.frame(
  session_id        = rep(paste0("s", 1:3), each = 6),
  code              = rep(c("Ask", "Reply", "Ask", "Reply", "Ask", "Reply"), 3),
  actor_type        = rep(c("Human", "Human", "AI", "AI", "Human", "AI"), 3),
  order_in_session  = rep(1:6, 3),
  stringsAsFactors  = FALSE
)

# Without disambiguate=, this errors:
try(build_htna(overlap, actor_type = "actor_type"))

# With disambiguate=, the two `Ask`s become distinct nodes:
net_dis <- build_htna(overlap, actor_type = "actor_type",
                      disambiguate = TRUE)
net_dis$nodes$label
```

## Splitting into cohorts with `group =`

Any of the three input forms can additionally be split into per-cohort
networks by passing `group =` — the result is an `htna_group`, one
`htna` network per level of the grouping column. Below we synthesise a
"phase" cohort variable on the combined frame and let `build_htna()`
build one network per phase.

```{r}
combined$phase <- ifelse(combined$session_id < median(combined$session_id),
                         "early", "late")

grp <- build_htna(combined, actor_type = "actor_type", group = "phase")
class(grp)
names(grp)
```

```{r, fig.height = 7}
plot_htna(grp)
```

## Cheat sheet

| Form | When to use |
|---|---|
| `list(Human = h, AI = a)` | Each actor's data is in its own frame |
| `df, actor_type = "actor_type"` | One combined frame, actor identity in a row-level column |
| `df, node_groups = list(Human = ..., AI = ...)` | One combined frame; partition declared as a named list of codes |
| `df, node_groups = data.frame(code = ..., actor_type = ...)` | One combined frame; partition declared as a tidy codebook |
| `+ actor = "user_id"` | Identify the individual actor that performed each event (forwarded to `Nestimate::build_network()`) |
| `+ disambiguate = TRUE` | Same code label appears in more than one actor type |
| `+ group = "phase"` | Build one network per cohort/phase |

All three input shapes are interchangeable; pick whichever matches the
shape your data already has.
