---
title: "Model assessment for htna networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model assessment for htna networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 8, fig.height = 5, dpi = 100
)
```



A heterogeneous transition network is a sample-based estimate of a
generative process. Each edge weight is a statistic computed from
the observed sequences, and as with any statistic the value carries
sampling variability. Interpretive claims drawn from a single
estimate — that an edge is strong, that one node is more central
than another, that a particular state dominates a phase of the
trajectory — depend on assumptions about that variability. Model
assessment is the procedure by which those assumptions are
quantified, before the network is used as evidence.

Four assessments are commonly applied to transition networks
(Saqr et al., 2025) and are exposed by the htna package as wrappers
around the `Nestimate` implementations:

1. **Split-half reliability**: estimates the agreement between two
   networks built from independent halves of the corpus.
2. **Edge-weight case-dropping stability**: estimates the proportion
   of cases that may be removed before the rank-ordering of edge
   weights deteriorates.
3. **Centrality stability**: estimates the proportion of cases that
   may be removed before node-level centrality rankings cease to be
   stable.
4. **Markov-order adequacy**: tests whether a first-order Markov
   model is sufficient to describe the observed transitions, or
   whether higher-order dependencies are required.

Each assessment is computed by a single function and visualised by
calling `plot()` on the result. The htna wrappers preserve the
actor partition (`$node_groups`, `$actor_levels`) on resampled
networks, so the assessment results remain compatible with downstream
htna-aware analysis.

# Data and baseline network

The example uses the bundled `human_ai` corpus -- Human + AI events
in a single long frame, tagged by an `actor_type` column (see
`?human_ai`). The default htna network is constructed under the
relative-probability scheme (`method = "relative"`); the
Markov-order test operates directly on the raw sequence data.

```{r load}
library(htna)
data(human_ai)

net <- build_htna(human_ai, actor_type = "actor_type")
```

# Split-half reliability

Split-half reliability quantifies the consistency of the network
estimate across random partitions of the corpus. At each iteration,
the sessions are partitioned uniformly at random into two halves; a
network is estimated on each half; and four discrepancy metrics are
computed between the two estimates: mean absolute deviation, median
absolute deviation, maximum absolute deviation, and the Pearson
correlation between the flattened edge-weight vectors. The procedure
is repeated `iter` times and the metrics are summarised across
iterations.

The Pearson correlation provides a one-number summary of similarity.
Values above 0.95 indicate near-identical estimates between halves;
values between 0.90 and 0.95 indicate substantively similar estimates
with minor sample-dependent variation; values below 0.85 indicate
that interpretation should be tempered.

```{r reliability}
rel <- reliability_htna(net, iter = 200L, seed = 1L)
rel
```

```{r reliability-plot}
plot(rel)
```

 

# Edge-weight case-dropping stability

Case-dropping stability refines reliability by quantifying how much
of the sample may be removed before the *ordering* of edge weights
changes. For each drop proportion in a fixed grid (default: 0.1 to
0.9 in steps of 0.1), a fraction of the sessions is removed at
random, the network is re-estimated, and the rank correlation
(Spearman by default; configurable via `method =`) between the
flattened edge-weight vector (off-diagonal by default; see
`include_diag`) of the reduced network and the original is recorded.
The procedure is repeated `iter` times per drop proportion.

The CS-coefficient (Epskamp et al., 2018) is defined as the maximum
drop proportion at which the rank correlation remains above
`threshold` (default 0.7) in at least `certainty` of the iterations
(default 0.95). A coefficient of 0.5 indicates that the edge ranking
is robust to removing half of the sample. Conventional cut-offs
(Epskamp et al., 2018) treat CS > 0.25 as acceptable and CS > 0.5
as preferred for inferential use.

```{r casedrop}
cd <- casedrop_reliability_htna(net, iter = 200L, seed = 1L)
cd
```

```{r casedrop-plot}
plot(cd)
```

The four-panel display shows the across-iteration mean of each
metric (mean / median / maximum absolute deviation, and rank
correlation) across drop proportions, with ribbons at ±1 SD across
iterations. The dashed reference line on the correlation panel
marks the threshold used to compute the CS-coefficient.

# Centrality stability

Reliability and case-dropping stability address the network as a
whole and the edge ranking, respectively. Centrality stability
addresses the node-level summaries that are commonly the target of
substantive interpretation. Many studies that report on transition
networks do not interpret individual edges; they interpret rankings
of nodes by centrality measures such as in-strength, out-strength,
or betweenness. Such rankings are only safely interpretable when
they are themselves stable under resampling.

`centrality_stability_htna()` applies the case-dropping procedure to
node-level centralities, computing each measure at every iteration
and quantifying the rank consistency of nodes across iterations.
The CS-coefficient is reported per measure.

```{r cstab}
cs <- centrality_stability_htna(
  net,
  measures = c("InStrength", "OutStrength", "Betweenness"),
  iter     = 200L,
  seed     = 1L
)
cs
```

```{r cstab-plot}
plot(cs)
```

In-strength and out-strength are typically more stable than
path-based measures because they aggregate weighted incoming and
outgoing edges directly. Betweenness depends on shortest paths and
is sensitive to small changes in edge weights that re-route many
paths simultaneously; consequently, its CS-coefficient is often
substantially lower than that of strength-based measures.
Interpretation by ranking should be limited to measures whose
CS-coefficient meets the conventional threshold.

By default `centrality_stability_htna()` evaluates only the three
measures that produce numerically identical results between cograph
and Nestimate (`InStrength`, `OutStrength`, `Betweenness`). The
remaining six htna centrality measures (closeness variants, RSP
betweenness, diffusion, weighted clustering) may be assessed by
supplying a custom `centrality_fn` argument.

# Markov-order adequacy

The preceding three assessments take the model — a first-order
Markov transition network — as a given and characterise its
sample-based variability. Markov-order adequacy addresses a more
fundamental question: whether the first-order assumption itself is
appropriate for the observed sequences.

`markov_order_test_htna()` tests the empirical transition structure
against orders 1, 2, ..., `max_order` via permutation. For each
candidate order *k*, the null hypothesis is that order-(*k*-1) is
sufficient; the alternative is that order-*k* captures dependencies
beyond what order-(*k*-1) explains. The smallest order whose test
fails to reject the null is reported as the optimal order.

The test operates on raw sequence data rather than on the htna
network object, since it requires the empirical sequences directly.

```{r markov-data}
# Restore per-session chronological order (the bundled dataset is
# sorted by project + step, which interleaves sessions), then take
# the AI-only events.
ai_only <- human_ai[order(human_ai$session_id,
                          human_ai$order_in_session), ]
ai_only <- ai_only[ai_only$actor_type == "AI", ]
seqs    <- split(ai_only$code, ai_only$session_id)
seqs    <- seqs[lengths(seqs) >= 3L]
length(seqs)
```

Sequences shorter than three events do not inform tests beyond
order-1 and are excluded.

```{r markov-test}
mo <- markov_order_test_htna(seqs, max_order = 3L, n_perm = 200L,
                             seed = 1L)
mo$test_table
mo$optimal_order
```

```{r markov-plot}
plot(mo)
```

If `optimal_order > 1`, the first-order network is misspecified for
the corpus, and a higher-order model is indicated. htna does not
implement higher-order networks directly; users who require them may
construct higher-order or memory-augmented models via
`Nestimate::build_hon()` or `Nestimate::build_mogen()`.

# Joint interpretation

The four assessments are complementary, addressing distinct levels
of the inferential chain:

| Assessment | Quantity assessed | Conventional pass criterion |
| --- | --- | --- |
| Split-half reliability | Whole-network estimate | mean correlation > 0.95 |
| Edge-weight case-drop | Edge-rank ordering | CS-coefficient > 0.5 |
| Centrality stability | Node-rank ordering | CS-coefficient > 0.5 per measure |
| Markov-order test | Model specification | optimal order = 1 |

A network that meets all four criteria supports interpretation at
every level — whole network, individual edges, node rankings, and
underlying model. A failure on any single criterion identifies the
level at which inferential claims should be qualified.

# References

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating
psychological networks and their accuracy: A tutorial paper.
*Behavior Research Methods*, 50(1), 195-212.

Saqr, M., et al. (2025). Transition network analysis: A novel
framework for modeling, visualizing, and identifying the temporal
patterns of learners and learning processes.

