---
title: "Multi-Group HTNA (3+ Actor Groups)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Group HTNA (3+ Actor Groups)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

HTNA is not limited to two actor groups. This vignette demonstrates how to
build and visualise a heterogeneous transition network with three groups.

## Simulating a three-actor interaction

We simulate a collaborative coding session with three actors -- a Student, an
AI assistant, and a System (e.g. IDE or automated tooling) -- each with four
distinct codes:

```{r}
library(htna)

set.seed(42)

student_codes <- c("Ask", "Confirm", "Revise", "Evaluate")
ai_codes      <- c("Suggest", "Explain", "Execute", "Clarify")
system_codes  <- c("Log", "Notify", "Compile", "Sync")
sessions      <- paste0("S", sprintf("%03d", 1:30))

make_events <- function(codes, sessions, offset) {
  rows <- lapply(sessions, function(sid) {
    n <- sample(15:25, 1)
    data.frame(
      session_id       = sid,
      code             = sample(codes, n, replace = TRUE),
      order_in_session = seq_len(n) * 3L - offset,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

student_data <- make_events(student_codes, sessions, 2L)
ai_data      <- make_events(ai_codes, sessions, 1L)
system_data  <- make_events(system_codes, sessions, 0L)
```

## Building a 3-group network

Pass all three groups as a named list:

```{r}
net3 <- build_htna(
  list(Student = student_data, AI = ai_data, System = system_data)
)
net3
```

The actor partition now contains three levels:

```{r}
table(net3$node_groups$group)
```

## Plotting

With three or more groups, `plot_htna()` automatically switches to a polygon
(triangle) layout. The three colours from the default HTNA palette are applied
in order:

```{r}
plot_htna(net3)
```

## Per-actor sequences

`sequence_plot_htna()` works the same way with three actors. With
`by = "state"` each row is one (session, actor) and is coloured by
the concrete code; with `by = "group"` cells are coloured by actor
(here Student / AI / System):

```{r, fig.height = 7}
sequence_plot_htna(net3, by = "state", type = "index")
```

```{r}
sequence_plot_htna(net3, by = "group", type = "distribution")
```

The `by = "state"` legend is split into one block per actor with the
actor name as a sub-title, so the reader can tell at a glance which
codes belong to which actor.

## Comparing two networks

Build a second network from the same actors (e.g. with different
session-level dynamics) and use `plot_htna_diff()` for the elementwise
difference. Positive differences are green, negative red:

```{r}
set.seed(7)
student_data2 <- make_events(student_codes, sessions, 2L)
ai_data2      <- make_events(ai_codes,      sessions, 1L)
system_data2  <- make_events(system_codes,  sessions, 0L)

net3b <- build_htna(
  list(Student = student_data2, AI = ai_data2, System = system_data2)
)

plot_htna_diff(net3, net3b)
```

`permutation()` provides the non-parametric significance test; pass the
result to `plot_htna_diff()` to surface only the edges that differ
significantly:

```{r}
perm3 <- permutation(net3, net3b, iter = 200)
plot_htna_diff(perm3)
```

```{r}
plot_htna_diff(perm3, show_nonsig = TRUE)
```

## Meta-paths across three actors

Meta-paths now span all three actor types. For example, you can search for
paths where the System mediates between Student and AI:

```{r}
extract_meta_paths(net3, schema = "Student->System->AI")
```

Or enumerate all length-3 meta-paths:

```{r}
extract_meta_paths(net3, length = 3)
```
