---
title: "plot_htna: 2-Group Parameter Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot_htna: 2-Group Parameter Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  out.width = "100%",
  dpi = 150
)
library(htna)
```

This vignette demonstrates every `plot_htna()` parameter for 2-group
(bipartite) layouts, so you can see at a glance what each option does.

## Setup

Start by loading the package and the example data shipped with Nestimate:

```{r}
library(htna)
data(human_long, ai_long, package = "Nestimate")
```

Build the heterogeneous transition network:

```{r}
m <- build_htna(list(Human = human_long, AI = ai_long))
m
```




## 1. Defaults

Calling `plot_htna()` with no extra arguments uses the htna defaults: circular
layout, 0.05 edge threshold, and the built-in colour palette.

```{r defaults}
plot_htna(m)
```

## 2. Layout

The `layout` parameter controls how the two groups are arranged in space.

### Bipartite (default for 2 groups)

Places each group in a vertical column, with cross-group edges drawn between
them. This is the classic bipartite layout.

```{r layout-bipartite}
plot_htna(m, layout = "bipartite")
```

### Circular

Distributes all nodes along the perimeter of a circle, alternating between
groups. Good for emphasising the overall connectivity pattern.

```{r layout-circular}
plot_htna(m, layout = "circular")
```

## 3. Orientation

The `orientation` parameter rotates or rearranges the bipartite layout. It
only applies when `layout = "bipartite"` (or `"auto"` with 2 groups).

### Vertical (default)

Groups are placed in two vertical columns, side by side.

```{r orient-vertical}
plot_htna(m, orientation = "vertical")
```

### Horizontal

Groups are stacked as two horizontal rows, one above the other.

```{r orient-horizontal}
plot_htna(m, orientation = "horizontal")
```

### Facing

Similar to vertical, but the two columns face each other with nodes aligned
by rank, making it easier to trace individual edges.

```{r orient-facing}
plot_htna(m, orientation = "facing")
```

### Circular orientation

Arranges each group along a semicircle, so both groups together form a full
circle.

```{r orient-circular}
plot_htna(m, orientation = "circular")
```

## 4. Group spacing

The `group_spacing` parameter controls the horizontal distance between the two
columns. Larger values push the groups further apart, which can help reduce
edge overlap in dense networks.

```{r spacing-narrow}
plot_htna(m, group_spacing = 1)
```

```{r spacing-wide}
plot_htna(m, group_spacing = 12)
```

## 5. Node spacing

The `node_spacing` parameter controls the vertical gap between nodes within a
group. Increase it when node labels overlap; decrease it for a more compact
plot.

```{r nodespace-tight}
plot_htna(m, node_spacing = 0.3)
```

```{r nodespace-loose}
plot_htna(m, node_spacing = 1.5)
```

## 6. Columns

The `columns` parameter splits each group into multiple sub-columns. This is
useful when a group has many nodes and a single column would be too tall.
A scalar applies to both groups; a length-2 vector sets each group
independently.

```{r columns-2}
plot_htna(m, columns = 2)
```

```{r columns-asymmetric}
plot_htna(m, columns = c(1, 2))
```

### Column spacing

When using multiple columns, `column_spacing` adjusts the horizontal gap
between sub-columns within a group.

```{r column-spacing}
plot_htna(m, columns = 2, column_spacing = 0.5)
```

## 7. Node ordering

By default (`use_list_order = TRUE`), nodes appear in the order they are found
in the data. Setting it to `FALSE` reorders nodes by edge weight, placing the
most connected nodes at the centre of each group.

```{r order-weight}
plot_htna(m, use_list_order = FALSE)
```

## 8. Curvature

The `curvature` parameter controls how much cross-group edges bend. `0`
produces straight lines, while higher values increase the arc. Useful for
reducing overlapping edges.

```{r curv-straight}
plot_htna(m, curvature = 0)
```

```{r curv-high}
plot_htna(m, curvature = 1)
```

## 9. Intra-group edge curvature

The `intra_curvature` parameter adds extra curvature to edges between nodes in
the same group, arcing them away from the opposite group. This visually
separates within-group from between-group transitions.

```{r intra-curv}
plot_htna(m, intra_curvature = 0.5)
```

## 10. Group colors

### Using `group_colors` (recommended)

A vector of colours, one per group. This is the recommended approach because
it scales to any number of groups.

```{r grp-colors}
plot_htna(m, group_colors = c("red", "purple"))
```

### Using legacy `group1_color` / `group2_color`

These parameters still work for two-group networks but do not generalise to
three or more groups.

```{r grp12-colors}
plot_htna(m, group1_color = "green", group2_color = "darkred")
```

## 11. Edge colors

By default, edge colours are derived (darkened) from the group colours of the
source node.

### Custom edge colors

Supply a vector of colours to override the automatic derivation.

```{r edge-colors}
plot_htna(m, edge_colors = c("red", "blue"))
```

### Disable group-based edge coloring

Set `edge_colors = FALSE` to use a single default colour for all edges.

```{r edge-false}
plot_htna(m, edge_colors = FALSE)
```

## 12. Group shapes

The `group_shapes` parameter assigns a different node shape to each group,
adding a second visual channel beyond colour. Common shapes include
`"circle"`, `"square"`, `"diamond"`, and `"triangle"`.

```{r grp-shapes}
plot_htna(m, group_shapes = c("diamond", "triangle"))
```

### Legacy `group1_shape` / `group2_shape`

```{r grp12-shapes}
plot_htna(m, group1_shape = "triangle", group2_shape = "diamond")
```

## 13. Jitter

Jitter shifts nodes horizontally based on their connectivity, breaking the
rigid column alignment. Nodes with stronger cross-group edges move closer to
the other group, giving a visual indication of coupling strength.

```{r jitter-true}
plot_htna(m, jitter = TRUE)
```

### Control jitter amount

A numeric value controls the maximum displacement.

```{r jitter-amount}
plot_htna(m, jitter = 0.5)
```

### Jitter both sides

By default, only the first group is jittered. Set `jitter_side = "both"` to
apply jitter to both groups.

```{r jitter-both}
plot_htna(m, jitter = TRUE, jitter_side = "both")
```

### Jitter second group only

```{r jitter-second}
plot_htna(m, jitter = TRUE, jitter_side = "second")
```

## 14. Legend

### No legend

```{r legend-false}
# plot_htna(m, legend = FALSE)
```

### Legend position

The `legend_position` parameter accepts standard R positions: `"bottomright"`
(default), `"bottomleft"`, `"topleft"`, `"topright"`.

```{r legend-topleft}
plot_htna(m, legend_position = "topleft")
```

```{r legend-topright}
plot_htna(m, legend_position = "topright")
```

## 15. Extend lines

Draw extension lines outward from each node, which can help connect labels to
nodes in dense plots. `TRUE` uses a default length; a numeric value sets the
length explicitly.

```{r extend-lines}
plot_htna(m, extend_lines = TRUE)
```

```{r extend-lines-length}
plot_htna(m, extend_lines = 0.3)
```

## 16. Layout margin

The `layout_margin` parameter adds padding around the plot area. Increase it
when labels are clipped at the edges.

```{r margin-tight}
plot_htna(m, layout_margin = 0)
```

```{r margin-wide}
plot_htna(m, layout_margin = 0.4)
```

## 17. Scale

The `scale` parameter adjusts all spacing proportionally, useful when
exporting at higher resolutions. A value of 2 doubles all distances.

```{r scale-2}
#plot_htna(m, scale = 2)
```

## 18. Label abbreviation

The `label_abbrev` parameter truncates node labels to a fixed number of
characters (numeric) or lets the function choose automatically (`"auto"`).
Helpful when code names are long.

```{r abbrev-3}
plot_htna(m, label_abbrev = 3)
```

```{r abbrev-auto}
plot_htna(m, label_abbrev = "auto")
```

## 19. Custom node metadata

Pass a data frame with a `label` column (for matching) and a `labels` column
(for display) to rename nodes in the plot without modifying the network.

```{r nodes-df}
nodes_df <- data.frame(
  label  = c("left", "right"),
  labels = paste0("N", 1:10)
)
plot_htna(m, nodes = nodes_df)
```

## 20. Angle spacing (circular/polygon layouts)

The `angle_spacing` parameter controls the empty space at corners in circular
and polygon layouts. Higher values create larger gaps at the vertices, visually
separating the groups.

```{r angle-spacing}
plot_htna(m, layout = "circular", angle_spacing = 0.3)
```

## 21. Passthrough arguments

Any additional arguments are forwarded to `cograph::splot()` via `...`. This
gives access to the full set of rendering options.

### Hide weak edges

The `threshold` parameter (passed to splot) hides edges with weights below the
given value.

```{r minimum}
plot_htna(m, threshold = 0.1)
```

### Hide edge labels

```{r no-edge-labels}
plot_htna(m, edge.labels = FALSE)
```

## 22. Combining parameters

Parameters can be freely combined to fine-tune the visualisation:

```{r combined}
plot_htna(m,
          group_colors = c("#E91E63", "#009688"),
          group_shapes = c("circle", "diamond"),
          group_spacing = 6,
          jitter = TRUE,
          jitter_side = "both",
          curvature = 0.6,
          threshold = 0.08,
          legend_position = "topright")
```
