---
title: "Sequence pattern comparison: early vs late Human-AI interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequence pattern comparison: early vs late Human-AI interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 10, fig.height = 5,
  fig.align = "center", out.width = "100%", dpi = 150
)
set.seed(20260413)
```

# Background

A transition network represents the joint distribution of single-step
state transitions, marginalising over higher-order structure. When
the analytic question concerns recurring multi-step patterns —
fixed-length subsequences (k-grams) such as `Command -> Specify ->
Execute` — single-step summaries are insufficient: they decompose the
pattern into independent steps and lose its identity as a unit.

`sequence_compare_htna()` operates at the k-gram level. For two or
more cohorts of sessions, the function enumerates every k-gram of
length within a specified range, computes the per-cohort frequency,
and tests whether the pattern's rate differs across cohorts under
either a chi-square or a permutation null model. Standardised
residuals identify which cohort each pattern characterises.

The example below applies the procedure to a within-session split:
the first half of every Human-AI session is labelled `"early"` and
the second half `"late"`. The analytic question is which multi-step
patterns characterise the early phase of an interaction and which
characterise the late phase.

# Data preparation

The two long-format datasets (`human_long`, `ai_long`) are combined
into a single corpus, with an `actor_type` column distinguishing the
two actors and a `half` column marking the within-session position.
Base R `ave()` is used for both the per-session count and the
per-session position; a single `ifelse()` writes the half label.

```{r data}
library(htna)
data(human_long, ai_long, package = "Nestimate")

human_long$actor_type <- "Human"
ai_long$actor_type    <- "AI"

combined <- rbind(human_long, ai_long)
combined <- combined[order(combined$session_id, combined$order_in_session), ]

n_per <- ave(combined$order_in_session, combined$session_id, FUN = length)
pos   <- ave(combined$order_in_session, combined$session_id, FUN = seq_along)
combined$half <- ifelse(pos <= n_per %/% 2, "early", "late")

table(combined$half)
```

# Constructing the grouped htna network

`build_htna()` accepts both `actor_type` (the within-session actor
partition) and `group` (the between-session cohort grouping)
simultaneously. The result is an object of class `htna_group`: a
named list of htna networks, one per cohort, each carrying the
actor partition. Cohort labels are read by `sequence_compare_htna()`
from the list names; no separate `group` argument is required at
the comparison step.

```{r build}
net_g <- build_htna(
  combined,
  actor_type = "actor_type",
  action     = "code",
  group      = "half"
)
class(net_g)
names(net_g)
```

# Pattern comparison

`sequence_compare_htna()` is invoked on the grouped network. Pattern
lengths 3 to 5 are scanned; only patterns occurring at least 25
times across the corpus are retained. The chi-square test is used
because the within-session split makes the two halves
non-independent (each session contributes to both halves), violating
the exchangeability assumption that the permutation test relies on.
False-discovery-rate correction (`adjust = "fdr"`) controls for
multiple testing across the candidate patterns.

```{r compare}
res <- sequence_compare_htna(
  net_g,
  sub      = 3:5,
  min_freq = 25L,
  test     = "chisq",
  adjust   = "fdr"
)
res
```

The discriminative patterns appear at the top of the patterns table,
ranked by test statistic.

```{r top-patterns}
head(res$patterns, 10)
```

## Interpreting the standardised residuals

For each pattern, the standardised residual is computed from a 2×2
contingency table of (this pattern present vs. all other patterns)
crossed with cohort:

$$\text{stdres}_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij} \cdot
(1 - r_i / N) \cdot (1 - c_j / N)}}$$

A positive residual on the `early` cohort indicates that the pattern
is over-represented in the first half of sessions; a positive
residual on the `late` cohort indicates over-representation in the
second half. By the standardised-residual interpretation,
$|z| > 1.96$ corresponds to $p < 0.05$ at the cell level, and
$|z| > 3$ provides strong evidence of over- or under-representation.

# Pyramid display

The pyramid display arranges the top patterns as back-to-back
horizontal bars, with each cohort on one side. Bar length encodes
the per-cohort frequency; the in-bar label gives the standardised
residual. The colour scale is shared across both sides so that
residuals are directly comparable.

```{r pyramid}
plot(res, style = "pyramid", show_residuals = TRUE)
```

The pyramid display requires exactly two cohorts and is the most
direct visual comparison when this condition is met.

# Heatmap display

The heatmap display generalises to any number of cohorts. Each
column is a cohort; each row is a pattern; cell colour encodes the
standardised residual under the same scale as the pyramid display.

```{r heatmap}
plot(res, style = "heatmap")
```

# Sorting by frequency

By default, patterns are ranked by the test statistic — the most
discriminative patterns appear first. Ranking by total occurrence
count is alternative when the analytic interest is on the most
common patterns regardless of their cohort difference.

```{r sort-frequency}
plot(res, style = "pyramid", sort = "frequency", show_residuals = TRUE)
```

# Choice of test statistic

The two test choices supported by `sequence_compare_htna()`
correspond to different assumptions about the source of the
sessions in each cohort.

- **Chi-square** (`test = "chisq"`) tests whether the observed
  per-pattern, per-cohort counts depart from the expected counts
  under independence between pattern and cohort. The test makes no
  assumption of independence between cohorts; it is appropriate when
  the same units (sessions, participants) contribute to multiple
  cohorts, as in the within-session split used here.
- **Permutation** (`test = "permutation"`) shuffles cohort labels at
  the session level and assumes exchangeability across sessions. It
  is appropriate when the cohorts comprise independent groups of
  sessions — for example, sessions from different projects or from
  different experimental conditions — and not when sessions
  contribute to both cohorts.

Selection of the test should follow from the design that produced
the cohort labels.

# Related vignettes

The companion `frequencies-and-mosaic` vignette covers single-step
transition residuals (`mosaic_plot_htna()`), which answer the
analogous question at the level of individual transitions. The
`sequence-plots` vignette covers descriptive sequence-level
visualisations. The `model-assessment` vignette covers reliability
and stability of the network estimate.
