---
title: "Sequence plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 9, fig.height = 5.5, dpi = 100
)
```

# Background

A transition network compresses a corpus of trajectories into a
single graph. The network representation supports the analysis of
*structure* — connectivity, centrality, and aggregate transition
weights — but does not retain the temporal ordering of events within
sessions. Analyses concerned with *dynamics* (the timing of states
within a session, alternation between actors, or differences between
early- and late-session behaviour) require complementary
visualisation of the sequences themselves.

`sequence_plot_htna()` provides this complementary view for
heterogeneous transition networks. The function renders each session
as a coloured strip, accommodating large numbers of sessions in a
single figure, and exposes two orthogonal arguments that determine
the visual encoding:

- **`type`** controls layout: `"index"` (default; one panel per
  actor, vertically stacked, each row a session), `"heatmap"` (a
  single carpet across both actors with a separator at the actor
  boundary), or `"distribution"` (a stacked-area summary across
  sessions).
- **`by`** controls colour semantics: `"state"` (default; each state
  code receives a distinct colour, suitable for examining within-actor
  vocabularies) or `"group"` (each actor receives a single colour
  applied to the combined session matrix, suitable for examining
  turn-taking).

The two arguments are independent. The remaining sections present
each of the six combinations and identify the question each is
suited to.

# Data and network

The example uses the `Human` and `AI` long-format datasets shipped
with `Nestimate`. The actor partition (Human / AI) is recorded as
node-level metadata on the resulting htna network and is consumed
automatically by every plot below.

```{r load}
library(htna)
data(human_long, ai_long, package = "Nestimate")

net <- build_htna(list(Human = human_long, AI = ai_long))
```

# Per-state colouring (`by = "state"`)

Per-state colouring assigns a distinct colour to each state code,
allowing the within-actor vocabulary to be inspected directly. The
two-actor case yields two non-overlapping palettes, since htna actor
vocabularies are typically disjoint by construction.

## Index layout

The index layout assigns one panel per actor and renders each
session as a horizontal strip within that panel. The horizontal axis
is time; the vertical axis enumerates sessions, sorted by similarity
under a longest-common-subsequence metric. Reading is by row: the
sequence of colours along a session strip indicates the order of
codes produced by that actor.

```{r}
sequence_plot_htna(net)
```

Vertically aligned colour blocks within a panel correspond to states
that occur in long uninterrupted runs across many sessions; these
visual signatures map directly to strong self-loops in the
underlying transition network.

## Heatmap layout

The heatmap layout combines the per-actor panels into a single
carpet, separated by a white band marking the actor boundary. This
view emphasises corpus-level differences between actors in session
length and uniformity.

```{r}
sequence_plot_htna(net, type = "heatmap")
```

The white separator does not encode data; it is a visual marker
distinguishing the two actor regions of the otherwise contiguous
matrix.

## Distribution layout

The distribution layout aggregates across sessions: at each
timestep, the proportion of sessions occupied by each state is
plotted as a stacked area. The result summarises the state
composition of the corpus as a function of session time.

```{r}
sequence_plot_htna(net, type = "distribution")
```

A grey "NA" band appears in this view because each per-actor
sequence is NA-padded at the timesteps when the other actor was
acting. The padding is a structural feature of the per-actor
extraction. Removing it normalises the proportions over present
states only:

```{r}
sequence_plot_htna(net, type = "distribution", na = FALSE)
```

The NA-suppressed view is the appropriate display when the analytic
question concerns the state mix of each actor; the default view is
appropriate when the question concerns when each actor is silent.

# Per-actor colouring (`by = "group"`)

Per-actor colouring collapses the palette to one colour per actor,
applied to the combined session matrix. Within-actor detail is lost
in exchange for a clean turn-taking view: the colour at each cell
identifies which actor was acting at that timestep.

## Index layout

```{r}
sequence_plot_htna(net, by = "group")
```

Sessions in which actors alternate frequently appear striped;
sessions in which one actor dominates appear as solid blocks.

## Heatmap layout

```{r}
sequence_plot_htna(net, type = "heatmap", by = "group")
```

Per-actor faceting is unnecessary under group colouring because each
cell carries actor identity through its colour. The single carpet
layout emphasises macro-level patterns in actor balance and session
length.

## Distribution layout

```{r}
sequence_plot_htna(net, type = "distribution", by = "group")
```

The combined session matrix is fully populated (every timestep has
an active actor), so no NA band appears. The plot displays the
proportion of sessions in which each actor was acting at each
timestep — a turn-taking summary across the corpus.

# Selection criteria

The six combinations address distinct analytic questions:

| Analytic question | Recommended view |
| --- | --- |
| Within-session ordering of codes per actor | `type = "index"`, `by = "state"` |
| Corpus-level differences in session length and uniformity | `type = "heatmap"`, `by = "state"` |
| Evolution of state composition over session time, per actor | `type = "distribution"`, `by = "state"`, `na = FALSE` |
| Turn-taking pattern within individual sessions | `type = "index"`, `by = "group"` |
| Corpus-level actor balance over session time | `type = "distribution"`, `by = "group"` |

The default — `sequence_plot_htna(net)` — is a general-purpose
diagnostic suitable for an initial inspection of any htna corpus.

# Related vignettes

The companion `frequencies-and-mosaic` vignette covers the
time-marginalised view of the same network (state mix and
chi-square mosaic). The `model-assessment` vignette covers
reliability and stability of the network estimate.
