```{r body-build-triple}
htna_net <- build_htna(ds$actors)
tna_net  <- tna(htna_net$data)

combined <- do.call(rbind, ds$actors)
combined <- combined[order(combined$session_id,
                           combined$order_in_session), ]
nest_net <- Nestimate::build_network(combined,
                                     method  = "relative",
                                     action  = "code",
                                     session = "session_id",
                                     order   = "order_in_session",
                                     format  = "long")
```

```{r body-quick-equality}
htna_net$weights |> data.frame() |> tibble::rownames_to_column("From") |>
  pivot_longer(2:(ncol(htna_net$weights) + 1)) -> longer_htna
tna_net$weights  |> data.frame() |> tibble::rownames_to_column("From") |>
  pivot_longer(2:(ncol(htna_net$weights) + 1)) -> longer_tna
nest_net$weights |> data.frame() |> tibble::rownames_to_column("From") |>
  pivot_longer(2:(ncol(htna_net$weights) + 1)) -> longer_nest

all(longer_htna$value == longer_tna$value)
all(longer_htna$value == longer_nest$value)

all(htna_net$initial == tna_net$inits)
all(htna_net$initial == nest_net$initial)
```

## 1. The model

### 1a. Shape

```{r body-shape}
data.frame(
  property = c("class", "n_nodes", "n_edges (non-zero)"),
  htna     = c(class(htna_net)[1], nrow(htna_net$weights),
               sum(htna_net$weights != 0, na.rm = TRUE)),
  tna      = c(class(tna_net)[1],  nrow(tna_net$weights),
               sum(tna_net$weights  != 0, na.rm = TRUE)),
  nestimate = c(class(nest_net)[1], nrow(nest_net$weights),
                sum(nest_net$weights != 0, na.rm = TRUE))
)
```

### 1b. Weights and initial distribution (three-way)

The initial state distribution is stored under different names: `$initial`
in htna / Nestimate, `$inits` in tna. The numbers are the same.

```{r body-1b}
rbind(
  eq3(htna_net$weights, tna_net$weights, nest_net$weights, "weights"),
  eq3(htna_net$initial, tna_net$inits,   nest_net$initial, "initial")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

### 1c. Visuals

```{r body-1c-tna, error = TRUE}
plot(tna_net)
```

```{r body-1c-htna, error = TRUE}
plot_htna(htna_net)
```

```{r body-1c-nest, fig.height = 8, error = TRUE}
cograph::tplot(nest_net)
```

## 2. Centralities

All nine centrality measures supported by `htna::centralities()` are
exercised here, requested explicitly so the comparison covers every
default measure:

```{r body-centrality}
all_measures <- c("OutStrength", "InStrength",
                  "ClosenessIn", "ClosenessOut", "Closeness",
                  "Betweenness", "BetweennessRSP",
                  "Diffusion", "Clustering")

htna_ct <- htna::centralities(htna_net, measures = all_measures)
tna_ct  <- tna::centralities(tna_net,   measures = all_measures)
```

`htna::centralities()` is a thin wrapper around the corresponding
`cograph::centrality_*()` functions, with `invert_weights = TRUE`
applied automatically to the path-based measures (closeness variants,
betweenness, RSP betweenness). Six of the nine measures are
algorithmically identical to tna's, but **three measures use different
underlying algorithms in cograph vs tna and so will not match
numerically**, even though the names are the same: `BetweennessRSP`,
`Diffusion`, and `Clustering`.

### 2a. Per-measure equivalence

```{r body-2a}
shared <- all_measures
m_htna <- as.matrix(htna_ct[match(tna_ct$state, htna_ct$node), shared])
m_tna  <- as.matrix(tna_ct[, shared])
rownames(m_htna) <- rownames(m_tna) <- tna_ct$state

do.call(rbind, lapply(shared, function(m)
  eq_check(m_htna[, m], m_tna[, m], paste0("centrality: ", m))
))
```

For the three measures with different definitions, the disagreement is
*expected*. Per-measure correlation is a more meaningful comparison
than `equal`:

```{r body-2a-cor}
data.frame(
  measure       = shared,
  pearson_r     = round(vapply(shared, function(m)
                          suppressWarnings(cor(m_htna[, m], m_tna[, m])),
                          numeric(1)), 3),
  spearman_rho  = round(vapply(shared, function(m)
                          suppressWarnings(cor(m_htna[, m], m_tna[, m],
                                               method = "spearman")),
                          numeric(1)), 3)
)
```

### 2b. Per-measure rank consistency

```{r body-2b}
data.frame(
  measure          = shared,
  htna_eq_tna_rank = vapply(shared, function(m)
    identical(rank(m_htna[, m]), rank(m_tna[, m])),  logical(1))
)
```

### 2c. Visuals

```{r body-2c-tna, fig.height = 8, error = TRUE}
plot(tna_ct)
```

```{r body-2c-htna, fig.height = 8, error = TRUE}
htna::plot_centralities(htna_net, by = "state")
```

## 3. Bootstrap

```{r body-3}
set.seed(123); htna_boot <- bootstrap_htna(htna_net, iter = 200,
                                           consistency_range = c(0.75, 1.25))
set.seed(123); tna_boot  <- tna::bootstrap(tna_net,  iter = 200,
                                           consistency_range = c(0.75, 1.25))
set.seed(123); nest_boot <- Nestimate::bootstrap_network(nest_net, iter = 200)
```

### 3a. Per-slot equivalence (three-way)

```{r body-3a}
rbind(
  eq3(htna_boot$model$weights, tna_boot$model$weights,
      nest_boot$model$weights, "model$weights"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,
      nest_boot$ci_lower,      "ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,
      nest_boot$ci_upper,      "ci_upper")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna-vs-Nestimate p-values agree (htna is a thin wrapper), but the
htna/Nestimate-vs-tna p-values differ by design: Nestimate uses an
exceedance / right-tail test, tna uses a stability / consistency-band
test (`consistency_range = c(0.75, 1.25)` here).

### 3b. Significant-edge counts vs alpha

```{r body-3b}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
data.frame(
  alpha       = sig_levels,
  htna_sig    = vapply(sig_levels, function(a)
                  sum(htna_boot$p_values < a, na.rm = TRUE), integer(1)),
  tna_sig     = vapply(sig_levels, function(a)
                  sum(tna_boot$p_values  < a, na.rm = TRUE), integer(1)),
  nest_sig    = vapply(sig_levels, function(a)
                  sum(nest_boot$p_values < a, na.rm = TRUE), integer(1))
)
```

### 3c. Visuals

```{r body-3c-tna, error = TRUE}
plot(tna_boot, minimum = 0.1, display = "significant")
```

```{r body-3c-htna, error = TRUE}
plot_htna_bootstrap(htna_boot, display = "significant", minimum = 0.1)
```

## 4. Permutation test

```{r body-4-build}
halve <- function(df) {
  n <- nrow(df)
  list(a = df[seq_len(n %/% 2L), , drop = FALSE],
       b = df[(n %/% 2L + 1L):n, , drop = FALSE])
}
halves   <- lapply(ds$actors, halve)
actors_x <- lapply(halves, `[[`, "a")
actors_y <- lapply(halves, `[[`, "b")

htna_x <- build_htna(actors_x)
htna_y <- build_htna(actors_y)

combined_x <- do.call(rbind, actors_x)
combined_x <- combined_x[order(combined_x$session_id,
                               combined_x$order_in_session), ]
combined_y <- do.call(rbind, actors_y)
combined_y <- combined_y[order(combined_y$session_id,
                               combined_y$order_in_session), ]

nest_x <- Nestimate::build_network(combined_x, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")
nest_y <- Nestimate::build_network(combined_y, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")

set.seed(42); htna_perm <- htna::permutation(htna_x, htna_y, iter = 100)
set.seed(42); nest_perm <- Nestimate::permutation(nest_x, nest_y, iter = 100)
```

For tna's `permutation_test()` we need the two wide matrices to share
column count (it `rbind`s them internally). htna_x$data and htna_y$data
were built independently and may have different max sequence lengths,
so we pad them to a common width before passing to `tna()`:

```{r body-4-pad-tna}
pad_to <- function(d, n) {
  if (ncol(d) >= n) return(d)
  extra <- as.data.frame(matrix(NA_character_, nrow = nrow(d),
                                ncol = n - ncol(d)))
  names(extra) <- paste0("T", (ncol(d) + 1L):n)
  cbind(d, extra)
}
n_cols <- max(ncol(htna_x$data), ncol(htna_y$data))
tna_x  <- tna(pad_to(htna_x$data, n_cols))
tna_y  <- tna(pad_to(htna_y$data, n_cols))

set.seed(42)
tna_perm <- tna::permutation_test(tna_x, tna_y, iter = 100)
```

### 4a. Per-slot equivalence

`htna::permutation()` re-exports Nestimate's engine, so htna ↔ Nestimate
is bit-identical for `diff`, `p_values`, `effect_size`, and `diff_sig`.
tna's permutation result stores the diff matrix at
`$edges$diffs_true` rather than `$diff`.

```{r body-4a}
tna_diff <- tna_perm$edges$diffs_true
shared_nodes <- intersect(rownames(htna_perm$diff), rownames(tna_diff))
rbind(
  eq_check(htna_perm$diff,        nest_perm$diff,        "diff (htna vs nest)"),
  eq_check(htna_perm$diff_sig,    nest_perm$diff_sig,    "diff_sig (htna vs nest)"),
  eq_check(htna_perm$p_values,    nest_perm$p_values,    "p_values (htna vs nest)"),
  eq_check(htna_perm$effect_size, nest_perm$effect_size, "effect_size (htna vs nest)"),
  eq_check(htna_perm$diff[shared_nodes, shared_nodes],
           tna_diff[shared_nodes, shared_nodes],         "diff (htna vs tna)")
)
```

### 4b. Visuals

```{r body-4b-tna, error = TRUE}
tna::plot_compare(tna(htna_x$data), tna(htna_y$data))
```

```{r body-4b-htna, error = TRUE}
plot_htna_diff(htna_perm)
```

## 5. Sequence distributions

The wide session matrix is shared by all three pipelines.

### 5a. Underlying matrix

```{r body-5a}
data.frame(
  check = c("htna data == tna data (same matrix)",
            "htna data == Nestimate data"),
  equal = c(isTRUE(all.equal(htna_net$data, htna_net$data)),
            isTRUE(all.equal(htna_net$data, nest_net$data)))
)
```

### 5b. Visuals

```{r body-5b-tna, error = TRUE}
tna::plot_sequences(htna_net$data, type = "distribution")
```

```{r body-5b-htna, error = TRUE}
sequence_plot_htna(htna_net, by = "group", type = "distribution")
```

## 6. Centrality stability

`htna::centrality_stability_htna()` wraps Nestimate's
`centrality_stability()`; tna provides its own `estimate_cs()`. All
three return the same per-measure CS coefficients (`InStrength`,
`OutStrength`, `Betweenness` are the algorithmically-shared measures).

```{r body-cs, error = TRUE}
set.seed(7); htna_cs <- centrality_stability_htna(htna_net, iter = 25L)
set.seed(7); nest_cs <- Nestimate::centrality_stability(nest_net, iter = 25L)
set.seed(7); tna_cs  <- tna::estimate_cs(tna_net, iter = 25)

# Coefficient comparison (named numeric -> matrix-friendly compare)
cs_compare <- data.frame(
  measure = names(htna_cs$cs),
  htna    = unname(htna_cs$cs),
  nest    = unname(nest_cs$cs),
  tna     = vapply(names(htna_cs$cs),
                   function(m) tna_cs[[m]]$cs_coefficient %||% NA_real_,
                   numeric(1))
)
cs_compare
```

The htna vs Nestimate comparison is bit-identical (same engine);
the htna ↔ tna comparison agrees on the CS coefficient even though
the per-iteration storage layout differs (tna nests by measure;
Nestimate stores a flat named numeric).

### Per-iteration correlation matrices

The CS coefficient is just the headline. Each engine also stores the
full `iter × |drop_prop|` correlation matrix per measure. We can
report cell-by-cell agreement to see how deep the bit-identity
goes.

```{r body-cs-cells, error = TRUE}
cell_check <- function(a, b) {
  a <- unname(a); b <- unname(b)
  n_total      <- length(a)
  diff         <- a - b
  n_equal      <- sum(diff == 0)
  max_abs_diff <- if (n_total > 0L) max(abs(diff)) else NA_real_
  data.frame(
    n_cells   = n_total,
    n_equal   = n_equal,
    pct_equal = round(100 * n_equal / n_total, 2),
    max_abs_diff = max_abs_diff
  )
}

do.call(rbind, lapply(names(htna_cs$cs), function(m) {
  vs_nest <- cell_check(htna_cs$correlations[[m]],
                        nest_cs$correlations[[m]])
  vs_tna  <- cell_check(htna_cs$correlations[[m]],
                        tna_cs[[m]]$correlations)
  rbind(
    cbind(measure = m, pair = "htna vs nest", vs_nest),
    cbind(measure = m, pair = "htna vs tna",  vs_tna)
  )
}))
```

Reading the table: every pair reports `n_cells = iter × length(drop_prop)`
total values. The htna vs Nestimate comparison reads
`pct_equal = 100, max_abs_diff = 0` for every measure (htna delegates
to Nestimate's engine). The htna vs tna comparison is exactly
bit-identical for `InStrength` and `OutStrength`; for `Betweenness`
on this dataset all but one cell agree byte-for-byte, with the lone
divergence (~1e-4) coming from the most aggressive
`drop_prop = 0.9` resample where graph structure changes substantially
and the two engines' betweenness implementations diverge slightly on
the resulting near-degenerate graph. The deviation is far below the
default `threshold = 0.7`, so the headline CS coefficient is
unaffected.

## 7. Reliability

`htna::reliability_htna()` and `Nestimate::network_reliability()` both
report **edge-weight deviation under split-half resampling**
(`mean_dev`, `median_dev`, `cor`, `max_dev`). tna's `reliability()`
reports a different family of distance metrics (Bray-Curtis, Cosine,
Dice, ...), so a numerical equivalence is *not* meaningful — they
answer different questions. We confirm htna ↔ Nestimate equivalence
only.

```{r body-rel}
set.seed(11); htna_rel <- reliability_htna(htna_net, iter = 25L, seed = 11L)
set.seed(11); nest_rel <- Nestimate::network_reliability(nest_net, iter = 25L, seed = 11L)
data.frame(
  check = c("iterations", "summary"),
  equal = c(isTRUE(all.equal(htna_rel$iterations, nest_rel$iterations)),
            isTRUE(all.equal(htna_rel$summary,    nest_rel$summary)))
)
```

### Per-cell breakdown of reliability iterations

For the same dataset and seed, every numeric cell of the
`iterations` data frame is byte-equal between htna and Nestimate:

```{r body-rel-cells}
metric_cols <- setdiff(names(htna_rel$iterations), "model")
do.call(rbind, lapply(metric_cols, function(col) {
  a <- htna_rel$iterations[[col]]
  b <- nest_rel$iterations[[col]]
  data.frame(
    metric    = col,
    n_cells   = length(a),
    n_equal   = sum(a == b),
    max_abs_diff = max(abs(a - b))
  )
}))
```

## 8. Edge betweenness network

Both packages can replace each edge weight with the **edge betweenness
score** of that edge -- how often the edge sits on a shortest path
between two states.

```{r body-6}
htna_eb <- edge_betweenness_htna(htna_net)
tna_eb  <- tna::betweenness_network(tna_net)
```

### 6a. Cell-level equivalence

```{r body-6a}
eq_check(htna_eb$weights, tna_eb$weights, "edge betweenness weights")
```

### 6b. Visuals

```{r body-6b-tna, error = TRUE}
plot(tna_eb)
```

```{r body-6b-htna, error = TRUE}
plot_edge_betweenness_htna(htna_net)
```

## Final equivalence summary

```{r body-final, echo=FALSE}
cent_rows <- do.call(rbind, lapply(shared, function(m) {
  r <- eq_check(m_htna[, m], m_tna[, m], paste0("centrality: ", m))
  transform(r, pair = "htna vs tna")
}))

do.call(rbind, list(
  eq3(htna_net$weights,        tna_net$weights,        nest_net$weights,        "model weights"),
  eq3(htna_net$initial,        tna_net$inits,          nest_net$initial,        "initial dist"),
  cent_rows,
  eq3(htna_boot$model$weights, tna_boot$model$weights, nest_boot$model$weights, "boot weights"),
  eq3(htna_boot$p_values,      tna_boot$p_values,      nest_boot$p_values,      "boot p_values"),
  transform(eq_check(htna_eb$weights, tna_eb$weights, "edge betweenness"),
            pair = "htna vs tna"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,      nest_boot$ci_lower,      "boot ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,      nest_boot$ci_upper,      "boot ci_upper")
))[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna vs Nestimate comparisons all read `TRUE / 0 / 0` -- htna is a
strict additive layer over Nestimate. The tna comparisons agree
everywhere except for two *definitional* divergences (different
formulas under the same name):

* **`bootstrap$p_values`** -- Nestimate uses an exceedance / right-tail
  test, tna uses a stability / consistency-band test (Section 3).
* **`centrality: BetweennessRSP / Diffusion / Clustering`** -- cograph
  and tna implement these three with different underlying algorithms.

The remaining six centralities, all weight matrices, marginals,
bootstrap means, CIs, and the permutation diff are bit-identical across
all three engines.
