---
title: "Equivalence with tna: near-deterministic transitions"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)
source("_equivalence-helpers.R")
```

25 sessions (length 12--18) on a 3+3 alphabet where each row of the
transition matrix concentrates ~90% of its mass on a single successor,
forming a 6-cycle: `h1 → a1 → h2 → a2 → h3 → a3 → h1`. The resulting
weight matrix is sparse -- most cells are tiny "leakage" probabilities
-- and centrality measures are nearly constant by symmetry of the
cycle. The article confirms the engines agree on the sparse pattern
and on the (sometimes constant) centrality values.

The pipeline below runs the same htna ↔ tna ↔ Nestimate three-way
equivalence checks as the
[main equivalence article](tna-equivalence.html).

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

ds <- ds_near_deterministic()
```

```{r child = "_equivalence-body.Rmd"}
```
