---
title: "Equivalence with tna: tiny dataset"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)
source("_equivalence-helpers.R")
```

The minimum-viable amount of data: only 6 sessions of length 5--7, on
a tiny 2+2 alphabet (`h1, h2` for Human, `a1, a2` for AI). Bootstrap
and permutation resamples will frequently be near-degenerate, so very
few edges (often none) clear the consistency-band test at modest
iteration counts. The point of this article is to confirm the engines
still agree cell-for-cell even under these conditions -- including
agreeing on which edges fail to be significant.

The pipeline below runs the same htna ↔ tna ↔ Nestimate three-way
equivalence checks as the
[main equivalence article](tna-equivalence.html).

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

ds <- ds_tiny()
```

```{r child = "_equivalence-body.Rmd"}
```
