---
title: "Equivalence with tna"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
source("_equivalence-helpers.R")
```

`htna` is built on the same engine as `Nestimate`, the engine that `tna`
shares for many quantities. To make the equivalence picture explicit
this article runs **three** parallel pipelines on the same data:

* **htna**: `build_htna(ds$actors)` -- wraps `Nestimate::build_network()`
  and stores the actor partition.
* **tna**: `tna(htna_net$data)` -- tna's own engine on the wide matrix.
* **Nestimate**: `Nestimate::build_network(combined_long, ...)` -- the
  Nestimate engine without the htna partition wrapper.

For each quantity we check **all three pairwise** comparisons and report
`equal / max_abs_diff / n_disagree`. If all three engines produce the
same numbers, every row reads `TRUE / 0 / 0`.

This article uses Nestimate's `human_long` + `ai_long` data. Sibling
articles repeat the same pipeline on diverse and pathological datasets:
synthetic balanced, synthetic wide, tiny, imbalanced, disconnected,
variable-length, near-deterministic, and `tna::group_regulation_long`.

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

ds <- ds_nestimate()
```

```{r child = "_equivalence-body.Rmd"}
```
