---
title: "Equivalence with tna: group_regulation_long"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)
source("_equivalence-helpers.R")
```

Real-world data: `tna::group_regulation_long` re-cast as a two-actor
partition. The action vocabulary is split into a **Cognitive** group
(`plan, monitor, discuss, synthesis, adapt`) and a **Social** group
(`cohesion, consensus, emotion, coregulate`). Sessions are grouped by
`(Group, Course, Actor)` and ordered by `Time`. This is the largest
dataset in the equivalence series and uses an empirical sequence-
length distribution rather than a synthetic one.

The pipeline below runs the same htna ↔ tna ↔ Nestimate three-way
equivalence checks as the
[main equivalence article](tna-equivalence.html).

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

ds <- ds_tna_group_regulation()
```

```{r child = "_equivalence-body.Rmd"}
```
