---
title: "Equivalence with tna: disconnected actors"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7, dpi = 150
)
source("_equivalence-helpers.R")
```

24 sessions in which the two actors **never transition between each
other**: 12 Human-only sessions cycle through `h1`-`h3`, and 12 AI-only
sessions cycle through `a1`-`a3`. The resulting transition matrix is
block-diagonal, the graph has two disconnected components, and
path-based centralities (closeness variants, betweenness) can produce
`NaN` or `Inf` for cross-component pairs. The point of this article is
that **both engines should produce the same `NaN` / `Inf` values** at
the same cells.

The pipeline below runs the same htna ↔ tna ↔ Nestimate three-way
equivalence checks as the
[main equivalence article](tna-equivalence.html).

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

ds <- ds_disconnected()
```

```{r child = "_equivalence-body.Rmd"}
```
