---
title: "Synthetic Datasets for HTNA Validation"
author: "htna"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(htna)
set.seed(2026)
```

This document exercises every documented input shape and feature of
`build_htna()` against small, hand-built synthetic datasets. Each
section ships a self-contained dataset, builds the network, prints a
quick sanity check, and runs one downstream wrapper to make sure the
actor partition and metadata round-trip correctly.

The cohort split used in the grouped vignette (Early vs. Late, computed
from `session_date`) is materialised here as a real `phase` column on
the dataset so it can be passed straight to `group = "phase"`.

## Helpers

A single helper that fabricates a long-format frame from a per-actor
transition matrix. Sessions vary in length so every fixture exercises
the order/session bookkeeping.

```{r helpers}
sim_sessions <- function(actors, alphabets, trans, sessions,
                         len_range = c(6L, 14L)) {
  rows <- list()
  for (sid in sessions) {
    n  <- sample(seq(len_range[1], len_range[2]), 1L)
    a  <- sample(actors, 1L)
    seq_codes  <- character(n)
    seq_actors <- character(n)
    seq_codes[1]  <- sample(alphabets[[a]], 1L)
    seq_actors[1] <- a
    for (i in seq_len(n - 1L)) {
      probs <- trans[[a]][seq_codes[i], ]
      nxt   <- sample(names(probs), 1L, prob = probs)
      a     <- attr(alphabets, "owner")[nxt]
      seq_codes[i + 1L]  <- nxt
      seq_actors[i + 1L] <- a
    }
    rows[[sid]] <- data.frame(
      session_id       = sid,
      code             = seq_codes,
      actor_type       = seq_actors,
      order_in_session = seq_len(n),
      stringsAsFactors = FALSE
    )
  }
  do.call(rbind, rows)
}

make_owner <- function(alphabets) {
  owner <- unlist(lapply(names(alphabets), function(a)
    setNames(rep(a, length(alphabets[[a]])), alphabets[[a]])))
  attr(alphabets, "owner") <- owner
  alphabets
}

uniform_trans <- function(full_alphabet) {
  matrix(1 / length(full_alphabet),
         nrow = length(full_alphabet),
         ncol = length(full_alphabet),
         dimnames = list(full_alphabet, full_alphabet))
}
```

## Dataset 1 — Combined frame with `actor_type` (canonical)

The most common shape: one long frame, one row-level `actor_type`
column, two actor types. Session lengths vary 6–14 events. A `phase`
column (Early vs. Late) is included so the same fixture can be reused
for grouped builds in section 7.

```{r ds1-build}
alpha1 <- make_owner(list(
  Human = c("Ask", "Specify", "Refine", "Check"),
  AI    = c("Plan", "Execute", "Report", "Repair")
))
full1  <- unlist(alpha1, use.names = FALSE)
trans1 <- list(Human = uniform_trans(full1),
               AI    = uniform_trans(full1))

ds1 <- sim_sessions(
  actors    = c("Human", "AI"),
  alphabets = alpha1,
  trans     = trans1,
  sessions  = sprintf("S%03d", 1:40)
)

# Bake the early/late split into the dataset (chronological order by
# session_id is enough since session ids are zero-padded).
sess_ids        <- sort(unique(ds1$session_id))
early_ids       <- sess_ids[seq_len(length(sess_ids) %/% 2L)]
ds1$phase       <- ifelse(ds1$session_id %in% early_ids, "Early", "Late")
ds1$phase       <- factor(ds1$phase, levels = c("Early", "Late"))

head(ds1)
table(ds1$actor_type, ds1$phase)
```

```{r ds1-net}
net1 <- build_htna(ds1, actor_type = "actor_type")
net1
table(net1$node_groups$group)
plot_htna(net1)
```

## Dataset 2 — Named list of per-actor frames

Splits the same generator into two frames; the list names become the
actor labels. Each frame keeps the canonical schema. This exercises
the "Form 1" branch of `build_htna()`.

```{r ds2}
ds2_human <- ds1[ds1$actor_type == "Human",
                 c("session_id", "code", "order_in_session")]
ds2_ai    <- ds1[ds1$actor_type == "AI",
                 c("session_id", "code", "order_in_session")]

net2 <- build_htna(list(Human = ds2_human, AI = ds2_ai))
identical(sort(net2$node_groups$node), sort(net1$node_groups$node))
plot_htna(net2)
```

## Dataset 3 — `node_groups` as a named list

A combined frame with no actor-type column; the partition is declared
directly. Useful when the codebook is authoritative and the long frame
is "plain".

```{r ds3}
ds3 <- ds1[, c("session_id", "code", "order_in_session")]

net3 <- build_htna(
  ds3,
  node_groups = list(Human = alpha1$Human, AI = alpha1$AI)
)
net3
```

## Dataset 4 — `node_groups` as a tidy codebook data frame

Same partition expressed as a two-column lookup. The `code` column
matches the `action` parameter; the other column carries the actor
type.

```{r ds4}
codebook4 <- data.frame(
  code       = c(alpha1$Human, alpha1$AI),
  actor_type = factor(
    c(rep("Human", length(alpha1$Human)),
      rep("AI",    length(alpha1$AI))),
    levels = c("Human", "AI")
  ),
  stringsAsFactors = FALSE
)

net4 <- build_htna(ds3, node_groups = codebook4)
levels(net4$node_groups$group)            # honours factor ordering
```

## Dataset 5 — Three-actor heterogeneous network

Stresses the multigroup path: Student / AI / System with disjoint
alphabets. `plot_htna()` should auto-switch to a polygon layout.

```{r ds5}
alpha5 <- make_owner(list(
  Student = c("Ask", "Confirm", "Revise", "Evaluate"),
  AI      = c("Suggest", "Explain", "Execute", "Clarify"),
  System  = c("Log", "Notify", "Compile", "Sync")
))
full5  <- unlist(alpha5, use.names = FALSE)
trans5 <- list(Student = uniform_trans(full5),
               AI      = uniform_trans(full5),
               System  = uniform_trans(full5))

ds5 <- sim_sessions(
  actors    = c("Student", "AI", "System"),
  alphabets = alpha5,
  trans     = trans5,
  sessions  = sprintf("M%03d", 1:25),
  len_range = c(12L, 20L)
)

net5 <- build_htna(ds5, actor_type = "actor_type")
table(net5$node_groups$group)
plot_htna(net5)
```

## Dataset 6 — Overlapping code labels with `disambiguate = TRUE`

Both actors share the labels `Ask` and `Reply`. Without
`disambiguate = TRUE`, `build_htna()` errors. With it, codes are
prefixed with the actor label so they become distinct nodes.

```{r ds6}
ds6 <- data.frame(
  session_id       = rep(sprintf("O%02d", 1:8), each = 6L),
  code             = rep(c("Ask", "Reply", "Ask", "Reply", "Ask", "Reply"), 8L),
  actor_type       = rep(c("Human", "Human", "AI", "AI", "Human", "AI"), 8L),
  order_in_session = rep(1:6, 8L),
  stringsAsFactors = FALSE
)

# Default behaviour: error on label collision.
err <- tryCatch(
  build_htna(ds6, actor_type = "actor_type"),
  error = function(e) conditionMessage(e)
)
err

net6 <- build_htna(ds6, actor_type = "actor_type", disambiguate = TRUE)
net6$nodes$label
```

## Dataset 7 — Grouped build (`group = "phase"`) using the baked-in field

Reuses `ds1`, which already carries `phase` as a factor with levels
`c("Early", "Late")`. The result is one network per cohort over a
shared alphabet, so it feeds straight into `plot_htna_diff()` and
`permutation_htna()`.

```{r ds7-build}
grp <- build_htna(ds1, actor_type = "actor_type", group = "phase")
class(grp)
names(grp)
```

```{r ds7-plot, fig.height=5}
op <- par(mfrow = c(1, 2))
plot_htna(grp)
par(op)
```

```{r ds7-diff}
plot_htna_diff(grp$Late, grp$Early)
```

## Dataset 8 — Individual `actor` IDs (forwarded to Nestimate)

Adds a per-event `user_id` distinct from `actor_type`. The same actor
*type* may be performed by multiple individuals; `actor =` forwards
that grain to `Nestimate::build_network()`.

```{r ds8}
ds8 <- ds1
ds8$user_id <- ifelse(
  ds8$actor_type == "Human",
  sample(c("h01", "h02", "h03"), nrow(ds8), replace = TRUE),
  sample(c("ai_a", "ai_b"),     nrow(ds8), replace = TRUE)
)
table(ds8$actor_type, ds8$user_id)

net8 <- build_htna(ds8, actor_type = "actor_type", actor = "user_id")
net8
```

## Dataset 9 — Custom column names (non-default `session` / `order` / `action`)

The schema columns are renamed so we can verify the `action =`,
`session =`, and `order =` parameters all flow through. Same data as
`ds1` under different aliases.

```{r ds9}
ds9 <- ds1
names(ds9)[names(ds9) == "session_id"]       <- "conversation"
names(ds9)[names(ds9) == "code"]             <- "move"
names(ds9)[names(ds9) == "order_in_session"] <- "turn"

net9 <- build_htna(
  ds9,
  actor_type = "actor_type",
  action     = "move",
  session    = "conversation",
  order      = "turn"
)
identical(sort(net9$node_groups$node), sort(net1$node_groups$node))
```

## Dataset 10 — Factor-level ordering controls legend order

`actor_type` is a factor with the *less common* level listed first.
`build_htna()` honours declared factor levels for ordering even when
the order of first appearance differs.

```{r ds10}
ds10 <- ds1
ds10$actor_type <- factor(ds10$actor_type, levels = c("AI", "Human"))

net10 <- build_htna(ds10, actor_type = "actor_type")
levels(net10$node_groups$group)   # AI first, Human second
```

## Dataset 11 — Highly variable session lengths

Some sessions are length 2 (the minimum that produces a transition),
some are length 30+. This stresses bootstrap and reliability resampling
on heterogeneous session sizes.

```{r ds11}
ds11 <- sim_sessions(
  actors    = c("Human", "AI"),
  alphabets = alpha1,
  trans     = trans1,
  sessions  = sprintf("V%03d", 1:50),
  len_range = c(2L, 30L)
)
range(table(ds11$session_id))     # min/max session length

net11  <- build_htna(ds11, actor_type = "actor_type")
boot11 <- bootstrap_htna(net11, iter = 100)
plot_htna_bootstrap(boot11, display = "significant")
```

## Dataset 12 — Strongly biased transitions (sanity check for diff/perm)

Two cohorts share the alphabet but differ in transition structure:
`Early` is uniform, `Late` strongly prefers `Human:Ask -> AI:Plan`.
`permutation_htna()` should flag that edge as significant.

```{r ds12}
alpha12 <- make_owner(list(
  Human = c("Ask", "Specify"),
  AI    = c("Plan", "Execute")
))
full12 <- unlist(alpha12, use.names = FALSE)

trans_early <- list(Human = uniform_trans(full12),
                    AI    = uniform_trans(full12))
trans_late  <- trans_early
trans_late$Human["Ask", ] <- c(0.05, 0.05, 0.80, 0.10)

ds12_early <- sim_sessions(c("Human", "AI"), alpha12, trans_early,
                           sprintf("E%03d", 1:30))
ds12_late  <- sim_sessions(c("Human", "AI"), alpha12, trans_late,
                           sprintf("L%03d", 1:30))
ds12_early$phase <- "Early"
ds12_late$phase  <- "Late"
ds12 <- rbind(ds12_early, ds12_late)
ds12$phase <- factor(ds12$phase, levels = c("Early", "Late"))

grp12  <- build_htna(ds12, actor_type = "actor_type", group = "phase")
perm12 <- permutation_htna(grp12$Late, grp12$Early, iter = 300)
plot_htna_diff(perm12)
```

## Cheat sheet of fixtures

| Fixture | Form / feature exercised                              |
|---------|-------------------------------------------------------|
| `ds1`   | Combined frame, `actor_type`, baked-in `phase`        |
| `ds2`   | Named list of per-actor frames                        |
| `ds3`   | `node_groups` as a named list                         |
| `ds4`   | `node_groups` as a tidy codebook                      |
| `ds5`   | Three-actor multigroup                                |
| `ds6`   | Code-label collision + `disambiguate = TRUE`          |
| `ds7`   | `group = "phase"` from the baked-in field             |
| `ds8`   | Individual `actor =` (user IDs)                       |
| `ds9`   | Custom `action` / `session` / `order` column names    |
| `ds10`  | Factor-level ordering controls legend order           |
| `ds11`  | Highly variable session lengths                       |
| `ds12`  | Biased transitions, used for diff / permutation       |
