---
title: "State frequencies and chi-square mosaics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State frequencies and chi-square mosaics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 8, 
  fig.height = 5, 
  dpi = 90
)
```

In this vignette, we look at additional summary functions and 
visualizations in `htna`. We will use two networks  constructed from the 
bundled `human_ai` corpus (Human + AI events tagged by `actor_type`; 
see `?human_ai`). The frequency family operates on either, but the mosaic 
family requires an integer-weighted network because the chi-square test acts on
counts; this is the network produced under `method = "frequency"`.

```{r load}
library(htna)
data(human_ai)

net      <- build_htna(human_ai, actor_type = "actor_type")
net_freq <- build_htna(human_ai, actor_type = "actor_type",
                       method = "frequency")
```

# Marginal state distributions

## Tabular summary

`frequencies_htna()` returns the per-actor marginal state
distribution as a data frame with one row per (`actor_type`, `state`)
pair.
Columns are: actor type (`group`), state, count, and within-network
proportion. The table is the data underlying every chart in this
section.

```{r}
frequencies_htna(net)
```

## Graphical summary: `plot_frequencies_htna()`

`plot_frequencies_htna()` renders the same marginal distribution in
three layouts. Each layout encodes the same data but is optimised
for a different reading task.

### Treemap (`view = "treemap"`, default)

Each panel corresponds to one actor; tile area within a panel
encodes the within-network proportion of the corresponding state.
The layout is space-efficient and allows the full state vocabulary
to be displayed simultaneously.

```{r}
plot_frequencies_htna(net)
```

 

### Combined bars (`view = "bars"`)

The bars layout collapses both actor types onto a single y-axis, sorts
the states by total count, and colours the bars by actor. This
layout is appropriate when the analytic task is direct numerical
comparison across the full vocabulary.

```{r}
plot_frequencies_htna(net, view = "bars")
```

The bars layout returns a ggplot object, permitting modification
through standard ggplot composition operators:

```{r}
plot_frequencies_htna(net, view = "bars") +
  ggplot2::labs(title = "Human vs AI: state counts")
```

### Per-actor faceted bars (`view = "facet"`)

The faceted layout assigns each actor its own panel with an
independent y-axis. This is appropriate when the actors differ
substantially in event volume, since a shared scale would compress
the lower-volume actor's bars beyond legibility.

```{r}
plot_frequencies_htna(net, view = "facet")
```



# Joint transition distribution: chi-square mosaic

`mosaic_plot_htna()` displays the joint distribution of `(source,
target)` transitions as a chi-square mosaic. Each cell of the
transition matrix is rendered as a rectangle whose area is
proportional to the joint share of that transition; cell colour
encodes the standardised residual against an independence model.
Cells with positive residuals (over-represented relative to
independence) are coloured blue; cells with negative residuals
(under-represented) are coloured red; cells whose observed value
matches the independence prediction are white.

The chi-square test requires integer counts, hence the requirement
for a frequency-method network.

## Default residuals (permutation-based)

The default residual estimator is permutation-based, with
`n_perm = 500` iterations. Permutation residuals are appropriate
when cell counts are sparse or when the chi-square asymptotic
approximation is not trusted; the trade-off is computation time.

```{r}
mosaic_plot_htna(net_freq, seed = 1L)
```

Cells with strong positive residuals identify transitions that
characterise the process beyond what would be expected by chance;
cells with strong negative residuals identify transitions that are
systematically suppressed.

## Asymptotic residuals (`residuals = "asymptotic"`)

The closed-form chi-square standardised residual estimator
(`chisq.test()$stdres`) is faster and is the convention used by
`tna` and `vcd`. It is appropriate when cell counts are large
enough for the asymptotic approximation to hold.

```{r}
mosaic_plot_htna(net_freq, residuals = "asymptotic")
```

For htna corpora of typical size (hundreds of sessions, hundreds
to thousands of transitions per cell), permutation and asymptotic
residuals agree closely.

## Colour-scale clipping (`range = c(-4, 4)`)

By default the colour scale is calibrated to the maximum absolute
residual in the matrix. A single extreme cell can therefore
desaturate the remainder of the chart. Clipping the range to a
fixed interval preserves contrast across the matrix and supports
comparison of mosaics across networks.

```{r}
mosaic_plot_htna(net_freq, range = c(-4, 4), seed = 1L)
```

A range of ±4 is conventional in mosaic displays; residuals beyond
the range saturate to the most intense colour.

## Axis-label rotation (`top_angle`, `left_angle`)

For matrices with long state names or large vocabularies, axis
labels may collide. The `top_angle` and `left_angle` arguments
rotate the labels for legibility.

```{r}
mosaic_plot_htna(net_freq, top_angle = 45, left_angle = 0, seed = 1L)
```

 
