---
title: "Grouped HTNA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped HTNA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 90
)
```

A common HTNA question is whether the same actor partition behaves
*differently* across cohorts -- e.g. early-phase versus late-phase
sessions of an interaction corpus; control vs. experimental groups; 
female vs. male, etc.. `build_htna()` supports this
directly by passing the session-level grouping column to the function. 
The result is  **one network per cohort**, each preserving the actor partition.
Every other function in `htna` recognises the grouped result and
iterates over cohorts.

## An early / late example

We use the bundled `human_ai` corpus (Human + AI codes from
`Nestimate::human_long` / `ai_long`, with a slightly simplified
alphabet -- see `?human_ai`). Sessions are split chronologically:
ordered by their first `session_date` (with `session_id` as a
deterministic tiebreak), the first half of the sessions form the
*Early* cohort and the rest form the *Late* cohort, reflected 
in the `phase` column.

```{r}
library(htna)
data(human_ai)
table(human_ai$phase)
```

## Building one network per cohort

`build_htna()` takes the long data frame, the actor-type column, and a
`group` argument naming the session-level cohort column:

```{r}
grp <- build_htna(
  human_ai,
  actor_type = "actor_type",
  group      = "phase"
)
length(grp)
names(grp)
```

The result is a named list of HTNA networks. Indexing one element gives
a regular HTNA network you can use exactly like any other:

```{r}
grp$Early
grp$Late
```

Because both cohorts were built from the same alphabet, both networks
have the **same nodes** -- which is exactly what `plot_htna_diff()`
and `permutation_htna()` need.

## Plotting per cohort

`plot_htna()` iterates over cohorts and draws one network per element.
It does not manage the layout grid -- wrap with `par(mfrow = ...)` if
you want a panel:

```{r, fig.height = 5}
op <- par(mfrow = c(1, 2))
plot_htna(grp)
par(op)
```

## Per-actor sequences per cohort

`sequence_plot_htna()` works the same way:

```{r, fig.height = 8}
sequence_plot_htna(grp,   type = "index")
```

## Centralities per cohort

`centralities_htna(grp)` returns one tidy data frame, with a leading
`group` column identifying the cohort:

```{r}
ct <- centralities_htna(grp)
head(ct, 10)
```

`plot_centralities()` faces a grid: rows are cohorts, columns are
measures, so cohort-level differences are easy to read at a glance:

```{r, fig.height = 9}
plot_centralities(grp, by = "group")
```

## Bootstrap per cohort

`bootstrap_htna(grp)` runs the bootstrap on each cohort and returns
a group of bootstrap results. `plot_htna_bootstrap()` iterates again:

```{r, fig.height = 5}
boots <- bootstrap_htna(grp, iter = 200)
op <- par(mfrow = c(1, 2))
plot_htna_bootstrap(boots, display = "significant")
par(op)
```

## Centrality stability per cohort

`centrality_stability_htna(grp)` runs the case-dropping centrality
stability check on each cohort and returns a named list of
`htna_stability` objects. Each cohort's `cs` reports the largest
session-drop proportion at which `InStrength`, `OutStrength`, and
`Betweenness` still correlate with the originals at the default
0.7 / 0.95 threshold:

```{r}
stab <- centrality_stability_htna(grp, iter = 100, seed = 1)
sapply(stab, function(s) s$cs)
```

The columns are cohorts, the rows are centrality measures. By the
Epskamp et al. (2018) convention, CS > 0.25 is acceptable and
CS > 0.5 is preferred for inferential use; the maximum reportable
value is the upper bound of the drop-proportion grid.

## Comparing the two cohorts

Both networks share an alphabet, so you can compute the elementwise
difference directly with `plot_htna_diff()`. Positive (Late > Early)
is green, negative (Early > Late) is red:

```{r}
plot_htna_diff(grp$Late, grp$Early)
```

`permutation_htna()` quantifies which differences are statistically
significant:

```{r}
perm <- permutation_htna(grp$Late, grp$Early, iter = 500)
plot_htna_diff(perm)
```

```{r}
plot_htna_diff(perm, show_nonsig = TRUE)
```

## Cohort-level pattern comparison

`permutation_htna()` tests edge-level differences between cohorts.
For *pattern*-level differences — recurring k-grams that
characterise one cohort over the other — see the [sequence
comparison vignette](sequence-comparison.html). The same grouped
network feeds directly into `sequence_compare_htna()`, which also
supports a `level = "type"` option that runs the comparison on
meta-paths (e.g. `Human -> AI -> Human`) rather than concrete
state codes.

## Indexing

Because the result of `build_htna(..., group = ...)` is a named list,
all standard list operations work: `grp[["Early"]]`,
`grp[c("Early", "Late")]`, `length(grp)`, `names(grp)`. Subsets keep
the grouped class so they pass straight back to the wrappers above.
