---
title: "Markov-order adequacy in heterogeneous transition networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Markov-order adequacy in heterogeneous transition networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 8, fig.height = 5, dpi = 90
)
```


A first-order Markov network represents the transition probability
between consecutive states under the assumption that the next state
depends only on the current state and not on any earlier ones. This
assumption is convenient — it reduces a temporal process to a square
transition matrix and supports the entire htna analytic stack — but
it is also empirical. For some corpora, the data demand a higher
order: the next state depends jointly on the current and on one or
more preceding states, and a first-order model misrepresents the
dependence structure.

`markov_order_test_htna()` quantifies that demand. For a fixed range
of candidate orders, it fits a separate Markov model at each order
to the empirical sequences, computes the log-likelihood, and tests
whether each successive increase in order yields a statistically
significant improvement over the lower order via a permutation test
on the deviance statistic. The reported optimal order is the largest
order at which the test rejects equivalence to its predecessor.

In a heterogeneous setting (where two or more actors interleave in
the same session) the question gains an additional dimension. The
optimal order computed on the *combined* sequences is a property of
the joint Human-AI process. The optimal orders computed on the
*per-actor* sequences (extracting only one actor's events from each
session) are properties of each actor's own process. The two are not
guaranteed to agree. When they disagree, the disagreement is
substantive: it locates the temporal dependence in one actor's
behaviour rather than in the joint dynamics.

This vignette runs the test under three slicings of the example
Human-AI corpus and compares the resulting optimal orders.

# Data and three slicings

The example uses the bundled `human_ai` corpus -- Human + AI events
in a single long frame, tagged by an `actor_type` column (see
`?human_ai`). Three sequence sets are constructed:

- **Human-only**: each session restricted to its Human events,
  preserving order.
- **AI-only**: each session restricted to its AI events,
  preserving order.
- **Combined**: the original interleaved Human + AI sequence per
  session.

Sessions shorter than three events do not inform tests beyond
order-1 and are excluded from each set.

```{r data}
library(htna)
data(human_ai)

# Within each session, restore chronological order (the bundled
# dataset is sorted by project + step [order_in_session],
# which interleaves sessions).

human_ai <- human_ai[order(human_ai$session_id,
                           human_ai$order_in_session), ]

is_h <- human_ai$actor_type == "Human"
human_seqs <- split(human_ai$code[is_h], human_ai$session_id[is_h])
human_seqs <- human_seqs[lengths(human_seqs) >= 3L]

is_a <- human_ai$actor_type == "AI"
ai_seqs <- split(human_ai$code[is_a], human_ai$session_id[is_a])
ai_seqs <- ai_seqs[lengths(ai_seqs) >= 3L]

combined_seqs <- split(human_ai$code, human_ai$session_id)
combined_seqs <- combined_seqs[lengths(combined_seqs) >= 3L]

c(human = length(human_seqs),
  ai    = length(ai_seqs),
  combined = length(combined_seqs))
```

# Human sequences

The Human-only test asks whether the temporal dependence among
Human codes extends beyond order-1. The hypothesis at each order
*k* is that order-(*k*-1) is sufficient against the alternative
that order-*k* captures additional structure.

```{r human}
mo_h <- markov_order_test_htna(
  human_seqs,
  max_order = 4L,
  n_perm    = 200L,
  seed      = 1L
)
mo_h$test_table
mo_h$optimal_order
```

```{r human-plot}
plot(mo_h)
```

# AI sequences

The AI-only test asks the same question of the AI codes
independently, on the same scale of `max_order` and number of
permutations. The result does not depend on the Human sequences and
is therefore directly comparable to the Human result.

```{r ai}
mo_a <- markov_order_test_htna(
  ai_seqs,
  max_order = 4L,
  n_perm    = 200L,
  seed      = 1L
)
mo_a$test_table
mo_a$optimal_order
```

```{r ai-plot}
plot(mo_a)
```

# Combined sequences

The combined test treats the interleaved Human-AI sequence as a
single process over the union of state vocabularies. Higher-order
dependence in this view can arise from within-actor structure
(captured by the per-actor tests) or from cross-actor dependence
(an AI code reliably following a particular Human code or vice
versa) that the per-actor tests cannot see.

```{r combined}
mo_c <- markov_order_test_htna(
  combined_seqs,
  max_order = 4L,
  n_perm    = 200L,
  seed      = 1L
)
mo_c$test_table
mo_c$optimal_order
```

```{r combined-plot}
plot(mo_c)
```

# Comparative interpretation

The three optimal orders, considered together, partition the
temporal dependence in the corpus across actors and across the
joint process. Four interpretive cases are worth distinguishing:

| Pattern | Interpretation |
| --- | --- |
| All three optimal orders ≈ 1 | First-order htna network is well-specified; both actors and their interaction are memoryless at the resolution of this test. |
| Combined > both per-actor | The joint process carries dependence that neither actor's behaviour explains in isolation. The temporal structure resides in *cross-actor coupling* — turn-taking, response patterns, action-reaction loops. |
| One per-actor > the other and ≥ combined | The actor with the higher order is the one with internal long-range dependence. Their behaviour at step *t* depends on multiple prior steps of *their own* history, regardless of what the other actor did. |
| Per-actor orders ≥ combined | The per-actor processes individually carry more memory than their interleaved joint process. This is unusual and typically indicates that the combined sequence dilutes within-actor structure by alternating actors. |

Empirically, on the example corpus the Human-only optimal order is
**`r mo_h$optimal_order`** and the AI-only optimal order is
**`r mo_a$optimal_order`**. The combined optimal order is
**`r mo_c$optimal_order`**. All three saturate at the upper bound
of the search (see the next section), so the test cannot distinguish
which process carries the deepest internal memory at this
`max_order` — only that none of them is consistent with a
first-order assumption. The default htna network is therefore a
structural simplification of these data, even if it remains
analytically useful for descriptive and relational claims.

# A note on `optimal_order` saturating at `max_order`

When all tested orders show significant improvement over their
predecessor — as in this example — the reported `optimal_order`
equals the upper bound of the search (`max_order`). This is not the
"true" optimal order; it is the largest order tested. To probe
whether higher orders would still improve fit, increase
`max_order`. The trade-off is computational: the number of
parameters at order *k* is approximately $|S|^{k+1}$, where
$|S|$ is the size of the state alphabet, so each step roughly
multiplies fitting cost by the alphabet size.

# Practical recipe

1. Run the test at `max_order = 3` initially as a coarse screen.
2. If the optimal order saturates at the upper bound, increase
   `max_order` until the test reports a non-significant step.
3. For heterogeneous corpora, compute the test on per-actor and
   combined sequences separately, and consult the comparative
   table above.
4. If `optimal_order = 1` for the combined process, the htna
   first-order network is well-specified for the corpus.
5. If `optimal_order > 1` for the combined process but
   `optimal_order = 1` for each actor in isolation, the
   higher-order structure resides in cross-actor coupling. The
   first-order htna network captures aggregate transition
   patterns but understates the temporal coordination between
   actors.
6. If `optimal_order > 1` per-actor, the first-order htna network
   is structurally simplifying the dependence. The next section
   surveys principled alternatives.

# Alternatives when the first-order assumption is rejected

htna is a focused subset of the Nestimate ecosystem and does not
implement higher-order or non-Markovian network models directly.
When `markov_order_test_htna()` indicates that the first-order
assumption is inadequate, the analysis proceeds by constructing
the appropriate alternative model in Nestimate or, for selected
variants, in `tna`. The choice of alternative is determined by the
mechanism through which the first-order assumption fails.

## Higher-order Markov networks

If the rejection of order-1 reflects genuine dependence of the
next state on multiple preceding states, a higher-order Markov
network is the appropriate generalisation. Several constructors
in Nestimate implement higher-order models under different
encodings of the memory:

| Function | Description |
| --- | --- |
| `Nestimate::build_hon()` | Higher-order network constructed by introducing explicit memory nodes for sequences whose transition probabilities deviate from the first-order baseline. Appropriate when the optimal order is 2 or 3 and the analysis benefits from an explicit memory representation in the graph. |
| `Nestimate::build_honem()` | Higher-order network with memory encoded on edges rather than as expanded nodes. Yields a graph of the same node cardinality as the first-order network, with the trade-off that memory information is no longer locally readable from individual nodes. |
| `Nestimate::build_hypa()` | Higher-order network restricted to memory effects that are significant against a hypergeometric null model. Reduces the risk of over-fitting that arises when high-order transitions are estimated from limited data. |
| `Nestimate::build_mogen()` | Multi-order generative model in which the order is selected per node rather than fixed globally. Appropriate when memory depth is heterogeneous across states. |

```{r, eval=FALSE}
hon <- Nestimate::build_hon(combined_seqs, max_order = 2L)
```

The actor partition carried by an htna network is not preserved on
higher-order objects, because the partition is defined on
single-state nodes whereas higher-order models extend the node set
with memory configurations that have no direct actor identity.
Downstream analyses on a higher-order network therefore proceed
without partition-aware visualisation.

## Markov mixture models

A second mechanism by which the first-order test may reject the
null is corpus heterogeneity: the sample comprises several distinct
first-order processes whose mixture exhibits apparent higher-order
structure even though no individual process does. The
`Nestimate::build_mmm()` constructor estimates a *k*-component
mixture of first-order chains and assigns each session
probabilistically to a component.

| Function | Description |
| --- | --- |
| `Nestimate::build_mmm()` | Mixture of *k* first-order Markov chains with EM-based estimation. Each session receives posterior probabilities of membership in each component. Diagnostic of mixture rather than higher-order structure: the per-session log-likelihood improvement under a higher-order model is concentrated in a subset of sessions rather than uniformly distributed across the corpus. |

```{r, eval=FALSE}
mmm <- Nestimate::build_mmm(combined_seqs, k = 3L)
```

## TNA-family alternatives

Some apparent rejections of order-1 reflect not a Markov-order
problem but a transition-aggregation problem: the empirical
transitions exist but are weighted, thresholded, or attended to in
ways that the first-order relative-probability scheme does not
express. The family of estimations re-defines the transition
matrix under alternative aggregation rules without invoking
higher-order memory.

| Method | Description |
| --- | --- |
| `method = "attention"` | Attention-weighted transition network. Each transition is weighted by an attention coefficient that decays with sequential distance, yielding a continuous interpolation between first-order and longer-range coupling. |
| `method = "frequency"` | Frequency-thresholded transition network. Transitions below a configurable count are removed before estimation, restricting the model to a structurally robust subgraph. Appropriate when the first-order test passes overall but the network is dominated by sparse, low-evidence edges. |
 
These methods address the modelling-specification question
("what transition matrix represents this corpus?") rather than
the order-adequacy question ("does this corpus require memory?").
They are therefore complementary to the higher-order constructors
above rather than substitutes for them.

## Diagnostic summary

| Diagnostic from `markov_order_test_htna()` | Indicated next step |
| --- | --- |
| Optimal order = 1 across all slicings | First-order htna network is well-specified; no model change is indicated. |
| Optimal order > 1 in the combined process only | Cross-actor coupling carries dependence absent from each actor in isolation. Either retain the first-order network as a descriptive aggregate, or estimate `Nestimate::build_hon()` to represent the cross-actor memory explicitly. |
| Optimal order > 1 within one or both actors | Within-actor memory. `Nestimate::build_hon()` or `Nestimate::build_honem()` per actor, or `Nestimate::build_mogen()` when memory depth varies across states. |
| Optimal order saturates at `max_order` for large `max_order` | Test the corpus-mixture hypothesis with `Nestimate::build_mmm()`; high apparent order may reflect heterogeneous components rather than long memory. |
| First-order assumption holds but edges are sparse and noisy | Aggregation-specification rather than order question: `build_htna(..., method = "frequency")` to restrict to structurally robust transitions. |

# References

Saqr, M., López-Pernas, S., Törmänen, T., Kaliisa, R., Misiejuk, K., & Tikka, S. (2025). Transition network analysis: A novel framework for modeling, visualizing, and identifying the temporal patterns of learners and learning processes. Proceedings of the 15th International Learning Analytics and Knowledge Conference, 351–361. https://doi.org/10.1145/3706468.3706513

Tutz, G. (2012). *Regression for categorical data.* Cambridge University Press. https://doi.org/10.1017/cbo9780511842061

