---
title: "Sequence pattern comparison: early vs late Human-AI interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequence pattern comparison: early vs late Human-AI interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 10, fig.height = 5,
  fig.align = "center", out.width = "100%", dpi = 90
)
set.seed(20260413)
```


A transition network represents the joint distribution of single-step
state transitions, marginalising over higher-order structure. When
the analytic question concerns recurring multi-step patterns —
fixed-length subsequences (k-grams) such as `Command -> Specify ->
Execute` — single-step summaries are insufficient: they decompose the
pattern into independent steps and lose its identity as a unit.

`sequence_compare_htna()` operates at the k-gram level. For two or
more cohorts of sessions, the function enumerates every k-gram of
length within a specified range (e.g. `Request -> Specify -> Execute`),
computes the per-cohort frequency, and tests whether the pattern's
rate differs across cohorts under either a chi-square or a permutation
null model. Standardised residuals identify which cohort each pattern
characterises.

The example below applies the procedure to a between-session split:
sessions are ranked chronologically by their first `session_date`
(ties broken on `session_id`), the first half of sessions are
labelled `"Early"`, and the rest `"Late"`. The analytic question is
which patterns characterise the early phase of the
corpus's lifetime and which characterise the late phase.  

# Data preparation

The bundled `human_ai` corpus already carries an `actor_type` column
distinguishing Human and AI events (see `?human_ai`). A
session-level `phase` column is added, marking each session as
`"Early"` or `"Late"` based on its rank in chronological order.

```{r data}
library(htna)
data(human_ai)

# Chronological session order; session_id is a deterministic
# tiebreak among sessions that started on the same date.
sess_start <- aggregate(session_date ~ session_id, data = human_ai,
                        FUN = min)
sess_start <- sess_start[order(sess_start$session_date,
                               sess_start$session_id), ]
half       <- nrow(sess_start) %/% 2L
early_sess <- sess_start$session_id[seq_len(half)]

human_ai$phase <- ifelse(human_ai$session_id %in% early_sess,
                         "Early", "Late")
human_ai$phase <- factor(human_ai$phase, levels = c("Early", "Late"))
table(human_ai$phase)
```

# Constructing the grouped htna network

`build_htna()` accepts both `actor_type` (the within-session actor
partition) and `group` (the between-session cohort grouping)
simultaneously. The result is an object of class `htna_group`: a
named list of htna networks, one per cohort, each carrying the
actor partition. Cohort labels are read by `sequence_compare_htna()`
from the list names; no separate `group` argument is required at
the comparison step.

```{r build}
net_g <- build_htna(
  human_ai,
  actor_type = "actor_type",
  group      = "phase"
)
class(net_g)
names(net_g)
```

# Pattern comparison

`sequence_compare_htna()` is invoked on the grouped network. Pattern
lengths 3 to 5 are scanned; only patterns occurring at least 25
times across the corpus are retained. The permutation test is used
because each session belongs to exactly one cohort, so the
exchangeability assumption that underlies the permutation null is
satisfied. False-discovery-rate correction (`adjust = "fdr"`)
controls for multiple testing across the candidate patterns.

```{r compare}
res <- sequence_compare_htna(
  net_g,
  sub      = 3:5,
  min_freq = 25L,
  test     = "permutation",
  adjust   = "fdr"
)
res
```

The discriminative patterns appear at the top of the patterns table,
ranked by test statistic.

```{r top-patterns}
head(res$patterns, 10)
```

## Interpreting the standardised residuals

For each pattern, the standardised residual is computed from a 2×2
contingency table of (this pattern present vs. all other patterns)
crossed with cohort:

$$\text{stdres}_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij} \cdot
(1 - r_i / N) \cdot (1 - c_j / N)}}$$


A positive residual on the `Early` cohort indicates that the pattern
is over-represented in sessions from the first half of the corpus's
lifetime; a positive residual on the `Late` cohort indicates
over-representation in sessions from the second half. By the
standardised-residual interpretation, $|z| > 1.96$ corresponds to
$p < 0.05$ at the cell level, and $|z| > 3$ provides strong
evidence of over- or under-representation.

# Pyramid display

The pyramid display arranges the top patterns as back-to-back
horizontal bars, with each cohort on one side. Bar length encodes
the per-cohort frequency; the in-bar label gives the standardised
residual. The colour scale is shared across both sides so that
residuals are directly comparable.

```{r pyramid}
plot(res, style = "pyramid", show_residuals = TRUE)
```

The pyramid display requires exactly two cohorts and is the most
direct visual comparison when this condition is met.

# Heatmap display

The heatmap display generalises to any number of cohorts. Each
column is a cohort; each row is a pattern; cell colour encodes the
standardised residual under the same scale as the pyramid display.

```{r heatmap}
plot(res, style = "heatmap")
```

# Sorting by frequency

By default, patterns are ranked by the test statistic — the most
discriminative patterns appear first. Ranking by total occurrence
count is alternative when the analytic interest is on the most
common patterns regardless of their cohort difference.

```{r sort-frequency}
plot(res, style = "pyramid", sort = "frequency", show_residuals = TRUE)
```

# Choice of test statistic

The two test choices supported by `sequence_compare_htna()`
correspond to different assumptions about the source of the
sessions in each cohort.

- **Permutation** (`test = "permutation"`) shuffles cohort labels at
  the session level and assumes exchangeability across sessions. It
  is appropriate when the cohorts comprise independent groups of
  sessions — for example, sessions from different time periods,
  projects, or experimental conditions — as in the chronological
  split used here.
- **Chi-square** (`test = "chisq"`) tests whether the observed
  per-pattern, per-cohort counts depart from the expected counts
  under independence between pattern and cohort. The test makes no
  assumption of independence between cohorts and is therefore
  appropriate when the same units contribute to multiple cohorts —
  for example, a within-session early/late split where each session
  contributes events to both halves.

Selection of the test should follow from the design that produced
the cohort labels.
