---
title: "Equivalence with tna"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
```

```{r helpers, include=FALSE}
# Pairwise cell-level equivalence summary used throughout the article.
eq_check <- function(a, b, label = "") {
  if (is.null(a) || is.null(b)) {
    return(data.frame(check = label, equal = NA, max_abs_diff = NA_real_,
                      n_disagree = NA_integer_, stringsAsFactors = FALSE))
  }
  diff <- as.numeric(a - b)
  data.frame(
    check         = label,
    equal         = isTRUE(all.equal(a, b, check.attributes = FALSE)),
    max_abs_diff  = max(abs(diff), na.rm = TRUE),
    n_disagree    = sum(diff != 0, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

# Three-way comparison: returns one row labelling each pairwise check.
eq3 <- function(htna, tna, nest, label = "") {
  rbind(
    transform(eq_check(htna, tna,  paste0(label, " | htna vs tna")),  pair = "htna vs tna"),
    transform(eq_check(htna, nest, paste0(label, " | htna vs nest")), pair = "htna vs nest"),
    transform(eq_check(tna,  nest, paste0(label, " | tna vs nest")),  pair = "tna vs nest")
  )
}
```

`htna` is built on the same engine as `Nestimate`, the engine that `tna`
shares for many quantities. To make the equivalence picture explicit
this article runs **three** parallel pipelines on the same data:

* **htna**: `build_htna(list(Human = ..., AI = ...))` -- wraps
  `Nestimate::build_network()` and stores the actor partition.
* **tna**: `tna(htna_net$data)` -- tna's own engine on the wide matrix.
* **Nestimate**: `Nestimate::build_network(combined_long, ...)` -- the
  Nestimate engine without the htna partition wrapper.

For each quantity we check **all three pairwise** comparisons and report
`equal / max_abs_diff / n_disagree`. If all three engines produce the
same numbers, every row reads `TRUE / 0 / 0`.

## Setup

```{r}
library(tna)
library(htna)

data(human_long, ai_long, package = "Nestimate")
```

```{r}
htna_net <- build_htna(list(Human = human_long, AI = ai_long))
tna_net  <- tna(htna_net$data)

combined <- rbind(human_long, ai_long)
combined <- combined[order(combined$session_id, combined$order_in_session), ]
nest_net <- Nestimate::build_network(combined,
                                     method  = "relative",
                                     action  = "code",
                                     session = "session_id",
                                     order   = "order_in_session",
                                     format  = "long")
```

## 1. The model

### 1a. Shape

```{r}
data.frame(
  property = c("class", "n_nodes", "n_edges (non-zero)"),
  htna     = c(class(htna_net)[1], nrow(htna_net$weights),
               sum(htna_net$weights != 0, na.rm = TRUE)),
  tna      = c(class(tna_net)[1],  nrow(tna_net$weights),
               sum(tna_net$weights  != 0, na.rm = TRUE)),
  nestimate = c(class(nest_net)[1], nrow(nest_net$weights),
                sum(nest_net$weights != 0, na.rm = TRUE))
)
```

### 1b. Weights and initial distribution (three-way)

The initial state distribution is stored under different names: `$initial`
in htna / Nestimate, `$inits` in tna. The numbers are the same.

```{r}
rbind(
  eq3(htna_net$weights, tna_net$weights, nest_net$weights, "weights"),
  eq3(htna_net$initial, tna_net$inits,   nest_net$initial, "initial")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

### 1c. Visuals

```{r}
plot(tna_net)
```

```{r}
plot_htna(htna_net)
```
```{r, fig.height = 8}
cograph::tplot(nest_net)
```
## 2. Centralities

```{r}
htna_ct <- htna::centralities(htna_net)
tna_ct  <- tna::centralities(tna_net)
nest_ct <- tna::centralities(tna(nest_net$data))   # tna engine on Nestimate's data
```

### 2a. Per-measure equivalence (three-way)

```{r}
shared <- intersect(setdiff(names(tna_ct),  c("state", "group")),
                    setdiff(names(htna_ct), c("node",  "actor")))

# Align row order to tna's `state`
m_htna <- as.matrix(htna_ct[match(tna_ct$state, htna_ct$node), shared])
m_tna  <- as.matrix(tna_ct[, shared])
m_nest <- as.matrix(nest_ct[match(tna_ct$state, nest_ct$state), shared])
rownames(m_htna) <- rownames(m_tna) <- rownames(m_nest) <- tna_ct$state

do.call(rbind, lapply(shared, function(m)
  eq3(m_htna[, m], m_tna[, m], m_nest[, m], paste0("centrality: ", m))
))[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

### 2b. Per-measure rank consistency

```{r}
data.frame(
  measure          = shared,
  htna_eq_tna_rank = vapply(shared, function(m)
    identical(rank(m_htna[, m]), rank(m_tna[, m])),  logical(1)),
  htna_eq_nest_rank = vapply(shared, function(m)
    identical(rank(m_htna[, m]), rank(m_nest[, m])), logical(1)),
  tna_eq_nest_rank  = vapply(shared, function(m)
    identical(rank(m_tna[, m]),  rank(m_nest[, m])), logical(1))
)
```

 

### 2c. Visuals

```{r, fig.height = 8}
plot(tna_ct)
```

```{r, fig.height = 8}
htna::plot_centralities(htna_net, by = "group")
```


## 3. Bootstrap

```{r}
set.seed(123); htna_boot <- bootstrap_htna(htna_net, iter = 200)
set.seed(123); tna_boot  <- tna::bootstrap(tna_net,  iter = 200)
set.seed(123); nest_boot <- Nestimate::bootstrap_network(nest_net, iter = 200)
```

### 3a. Per-slot equivalence (three-way)

```{r}
rbind(
  eq3(htna_boot$model$weights, tna_boot$model$weights,
      nest_boot$model$weights, "model$weights"),
  eq3(htna_boot$p_values,      tna_boot$p_values,
      nest_boot$p_values,      "p_values"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,
      nest_boot$ci_lower,      "ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,
      nest_boot$ci_upper,      "ci_upper")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna-vs-Nestimate p-values agree (htna is a thin wrapper), but the
htna/Nestimate-vs-tna p-values differ. **This is a definitional
difference, not a bug.** Both engines use the formula
`(count + 1) / (iter + 1)`, but they count *different* events:

* **Nestimate** (and so htna): `count` is how often a resampled edge
  weight **exceeds** the observed weight -- a tail probability under
  the resampling distribution. Structural zeros get
  `p = 1 / (iter + 1) ≈ 0.005`.
* **tna**: `count` is how often a resampled weight falls **outside
  ±25%** of the observed weight (`consistency_range = c(0.75, 1.25)`)
  -- a stability test. Structural zeros have a band collapsed to a
  point, so every resample falls outside and `p = 1`.

The two tests answer different questions:

* "Is the observed weight surprisingly large under resampling?" (Nestimate)
* "Is the observed weight stable to within ±25% under resampling?" (tna)

The bootstrap *resamples themselves* are bit-identical -- which is why
`model$weights`, `ci_lower`, `ci_upper` all match -- but the test
statistic differs.

### 3b. Significant-edge counts vs alpha

```{r}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
data.frame(
  alpha       = sig_levels,
  htna_sig    = vapply(sig_levels, function(a) sum(htna_boot$p_values < a, na.rm = TRUE), integer(1)),
  tna_sig     = vapply(sig_levels, function(a) sum(tna_boot$p_values  < a, na.rm = TRUE), integer(1)),
  nest_sig    = vapply(sig_levels, function(a) sum(nest_boot$p_values < a, na.rm = TRUE), integer(1))
)
```

### 3c. Visuals

```{r}
plot(tna_boot)
```

```{r}
plot_htna_bootstrap(htna_boot, display = "significant")
```

## 4. Permutation test

```{r}
n_h <- nrow(human_long); n_a <- nrow(ai_long)

htna_x <- build_htna(list(Human = human_long[1:(n_h %/% 2), ],
                          AI    = ai_long[1:(n_a %/% 2), ]))
htna_y <- build_htna(list(Human = human_long[(n_h %/% 2 + 1):n_h, ],
                          AI    = ai_long[(n_a %/% 2 + 1):n_a, ]))

# tna's permutation_test requires the same number of columns; build the
# matched Nestimate networks via build_network so they share node order.
combined_x <- rbind(human_long[1:(n_h %/% 2), ],
                    ai_long[1:(n_a %/% 2), ])
combined_x <- combined_x[order(combined_x$session_id,
                               combined_x$order_in_session), ]
combined_y <- rbind(human_long[(n_h %/% 2 + 1):n_h, ],
                    ai_long[(n_a %/% 2 + 1):n_a, ])
combined_y <- combined_y[order(combined_y$session_id,
                               combined_y$order_in_session), ]

nest_x <- Nestimate::build_network(combined_x, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")
nest_y <- Nestimate::build_network(combined_y, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")

set.seed(42); htna_perm <- htna::permutation(htna_x, htna_y, iter = 200)
set.seed(42); nest_perm <- Nestimate::permutation(nest_x, nest_y, iter = 200)
```

### 4a. Per-slot equivalence (htna vs Nestimate)

`htna::permutation()` is a direct re-export of
`Nestimate::permutation()`, so the three-way picture collapses to one
non-trivial pair.

```{r}
rbind(
  eq_check(htna_perm$diff,        nest_perm$diff,        "diff"),
  eq_check(htna_perm$diff_sig,    nest_perm$diff_sig,    "diff_sig"),
  eq_check(htna_perm$p_values,    nest_perm$p_values,    "p_values"),
  eq_check(htna_perm$effect_size, nest_perm$effect_size, "effect_size")
)
```

### 4b. Significant-edge counts

```{r}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
data.frame(
  alpha     = sig_levels,
  htna_sig  = vapply(sig_levels, function(a) sum(htna_perm$p_values < a), integer(1)),
  nest_sig  = vapply(sig_levels, function(a) sum(nest_perm$p_values < a), integer(1))
)
```

### 4c. Top-5 strongest signed differences

```{r}
strongest <- function(perm) {
  d <- perm$diff
  idx <- which(d != 0 & perm$p_values < 0.05, arr.ind = TRUE)
  scores <- d[idx]
  ord <- order(abs(scores), decreasing = TRUE)[1:min(5, length(scores))]
  data.frame(
    from = rownames(d)[idx[ord, 1]],
    to   = colnames(d)[idx[ord, 2]],
    diff = round(scores[ord], 3),
    p    = signif(perm$p_values[idx[ord, , drop = FALSE]], 3),
    stringsAsFactors = FALSE
  )
}
strongest(htna_perm)
```

### 4d. Visuals

```{r}
tna::plot_compare(tna(htna_x$data), tna(htna_y$data))
```

```{r}
plot_htna_diff(htna_perm)
```

## 5. Sequence distributions

The wide session matrix is shared by all three pipelines.

### 5a. Underlying matrix

```{r}
data.frame(
  check        = c("htna data == tna data (same matrix)",
                   "htna data == Nestimate data"),
  equal        = c(isTRUE(all.equal(htna_net$data, htna_net$data)),
                   isTRUE(all.equal(htna_net$data, nest_net$data)))
)
```

### 5b. Per-time-step actor activity

```{r}
type_map <- setNames(as.character(htna_net$node_groups$group),
                     as.character(htna_net$node_groups$node))
state_mat <- as.matrix(htna_net$data)
active_t  <- function(t) {
  v <- state_mat[, t]; v <- v[!is.na(v)]
  data.frame(time = t, n_active = length(v),
             Human = sum(unname(type_map[v]) == "Human"),
             AI    = sum(unname(type_map[v]) == "AI"))
}
do.call(rbind, lapply(c(1, 5, 10, 20, 50, 100), active_t))
```

### 5c. Visuals

```{r}
tna::plot_sequences(htna_net$data, type = "distribution")
```

```{r}
sequence_plot_htna(htna_net, by = "group", type = "distribution")
```

## Final equivalence summary

```{r echo=FALSE}
do.call(rbind, list(
  eq3(htna_net$weights,        tna_net$weights,        nest_net$weights,        "model weights"),
  eq3(htna_net$initial,        tna_net$inits,          nest_net$initial,        "initial dist"),
  do.call(rbind, lapply(shared, function(m)
            eq3(m_htna[, m], m_tna[, m], m_nest[, m], paste0("centrality: ", m)))),
  eq3(htna_boot$model$weights, tna_boot$model$weights, nest_boot$model$weights, "boot weights"),
  eq3(htna_boot$p_values,      tna_boot$p_values,      nest_boot$p_values,      "boot p_values"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,      nest_boot$ci_lower,      "boot ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,      nest_boot$ci_upper,      "boot ci_upper")
))[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna vs Nestimate comparisons all read `TRUE / 0 / 0` -- htna is a
strict additive layer over Nestimate. The tna comparisons agree
everywhere except in the bootstrap p-values, where the two engines run
*different statistical tests* under the same name (an exceedance test
in Nestimate, a stability test in tna -- see Section 3 for details).
