---
title: "Equivalence with tna"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 10,
  fig.height = 7,
  dpi = 150
)
```

```{r helpers, include=FALSE}
# Pairwise cell-level equivalence summary used throughout the article.
eq_check <- function(a, b, label = "") {
  if (is.null(a) || is.null(b)) {
    return(data.frame(check = label, equal = NA, max_abs_diff = NA_real_,
                      n_disagree = NA_integer_, stringsAsFactors = FALSE))
  }
  diff <- as.numeric(a - b)
  data.frame(
    check         = label,
    equal         = isTRUE(all.equal(a, b, check.attributes = FALSE)),
    max_abs_diff  = max(abs(diff), na.rm = TRUE),
    n_disagree    = sum(diff != 0, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

# Three-way comparison: returns one row labelling each pairwise check.
eq3 <- function(htna, tna, nest, label = "") {
  rbind(
    transform(eq_check(htna, tna,  paste0(label, " | htna vs tna")),  pair = "htna vs tna"),
    transform(eq_check(htna, nest, paste0(label, " | htna vs nest")), pair = "htna vs nest"),
    transform(eq_check(tna,  nest, paste0(label, " | tna vs nest")),  pair = "tna vs nest")
  )
}
```

`htna` is built on the same engine as `Nestimate`, the engine that `tna`
shares for many quantities. To make the equivalence picture explicit
this article runs **three** parallel pipelines on the same data:

* **htna**: `build_htna(list(Human = ..., AI = ...))` -- wraps
  `Nestimate::build_network()` and stores the actor partition.
* **tna**: `tna(htna_net$data)` -- tna's own engine on the wide matrix.
* **Nestimate**: `Nestimate::build_network(combined_long, ...)` -- the
  Nestimate engine without the htna partition wrapper.

For each quantity we check **all three pairwise** comparisons and report
`equal / max_abs_diff / n_disagree`. If all three engines produce the
same numbers, every row reads `TRUE / 0 / 0`.

## Setup

```{r}
library(tna)
library(htna)
library(tidyr)
library(dplyr)
library(tibble)

data(human_long, ai_long, package = "Nestimate")
```

```{r}
htna_net <- build_htna(list(Human = human_long, AI = ai_long))
tna_net  <- tna(htna_net$data)

combined <- rbind(human_long, ai_long)
combined <- combined[order(combined$session_id, combined$order_in_session), ]
nest_net <- Nestimate::build_network(combined,
                                     method  = "relative",
                                     action  = "code",
                                     session = "session_id",
                                     order   = "order_in_session",
                                     format  = "long")
```


```{r}
htna_net$weights |> data.frame() |> tibble::rownames_to_column("From") |> pivot_longer(2:(ncol(htna_net$weights)+1)) -> longer_htna
tna_net$weights |> data.frame() |> tibble::rownames_to_column("From") |> pivot_longer(2:(ncol(htna_net$weights)+1)) -> longer_tna
nest_net$weights |> data.frame() |> tibble::rownames_to_column("From") |> pivot_longer(2:(ncol(htna_net$weights)+1))  -> longer_nest

all(longer_htna$value == longer_tna$value)
all(longer_htna$value == longer_nest$value)

all(htna_net$initial == tna_net$inits)
all(htna_net$initial == nest_net$inits)
```

## 1. The model

### 1a. Shape

```{r}
data.frame(
  property = c("class", "n_nodes", "n_edges (non-zero)"),
  htna     = c(class(htna_net)[1], nrow(htna_net$weights),
               sum(htna_net$weights != 0, na.rm = TRUE)),
  tna      = c(class(tna_net)[1],  nrow(tna_net$weights),
               sum(tna_net$weights  != 0, na.rm = TRUE)),
  nestimate = c(class(nest_net)[1], nrow(nest_net$weights),
                sum(nest_net$weights != 0, na.rm = TRUE))
)
```

### 1b. Weights and initial distribution (three-way)

The initial state distribution is stored under different names: `$initial`
in htna / Nestimate, `$inits` in tna. The numbers are the same.

```{r}
rbind(
  eq3(htna_net$weights, tna_net$weights, nest_net$weights, "weights"),
  eq3(htna_net$initial, tna_net$inits,   nest_net$initial, "initial")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

### 1c. Visuals

```{r}
plot(tna_net)
```

```{r}
plot_htna(htna_net)
```
```{r, fig.height = 8}
cograph::tplot(nest_net)
```
## 2. Centralities

All nine centrality measures supported by `htna::centralities()` are
exercised here, requested explicitly so the comparison covers every
default measure:

```{r}
all_measures <- c("OutStrength", "InStrength",
                  "ClosenessIn", "ClosenessOut", "Closeness",
                  "Betweenness", "BetweennessRSP",
                  "Diffusion", "Clustering")

htna_ct <- htna::centralities(htna_net, measures = all_measures)
tna_ct  <- tna::centralities(tna_net,   measures = all_measures)
```

`htna::centralities()` is now a thin wrapper around the corresponding
`cograph::centrality_*()` functions, with `invert_weights = TRUE`
applied automatically to the path-based measures (closeness variants,
betweenness, RSP betweenness). Six of the nine measures are
algorithmically identical to tna's, but **three measures use different
underlying algorithms in cograph vs tna and so will not match
numerically**, even though the names are the same:

| Measure          | htna (cograph)                                | tna                                           | Cell-equal? |
|------------------|-----------------------------------------------|-----------------------------------------------|-------------|
| `OutStrength`    | `centrality_outstrength`                      | `igraph::strength(mode = "out")`              | yes         |
| `InStrength`     | `centrality_instrength`                       | `igraph::strength(mode = "in")`               | yes         |
| `ClosenessIn`    | `centrality_incloseness` (inverted)           | `igraph::closeness(mode = "in")`              | yes         |
| `ClosenessOut`   | `centrality_outcloseness` (inverted)          | `igraph::closeness(mode = "out")`             | yes         |
| `Closeness`      | `centrality_closeness` (inverted)             | `igraph::closeness(mode = "all")`             | yes         |
| `Betweenness`    | `centrality_betweenness` (inverted)           | `igraph::betweenness`                         | yes         |
| `BetweennessRSP` | `centrality_current_flow_betweenness`         | RSP betweenness (`Z = (I−P·exp(−βC))⁻¹` form) | **no** (different algorithm) |
| `Diffusion`      | `centrality_diffusion`                        | `Σₖ Pᵏ` then row sums                         | **no** (different algorithm) |
| `Clustering`     | `centrality_transitivity`                     | weighted clustering on `M + Mᵀ`               | **no** (different algorithm) |

### 2a. Per-measure equivalence

```{r}
shared <- all_measures

# Align row order to tna's `state`
m_htna <- as.matrix(htna_ct[match(tna_ct$state, htna_ct$node), shared])
m_tna  <- as.matrix(tna_ct[, shared])
rownames(m_htna) <- rownames(m_tna) <- tna_ct$state

do.call(rbind, lapply(shared, function(m)
  eq_check(m_htna[, m], m_tna[, m], paste0("centrality: ", m))
))
```

For the three measures with different definitions, the disagreement is
*expected* — the names are shared but the formulas are not. Per-measure
correlation is a more meaningful comparison than `equal`:

```{r}
data.frame(
  measure       = shared,
  pearson_r     = round(vapply(shared, function(m)
                          cor(m_htna[, m], m_tna[, m]), numeric(1)), 3),
  spearman_rho  = round(vapply(shared, function(m)
                          cor(m_htna[, m], m_tna[, m],
                              method = "spearman"), numeric(1)), 3)
)
```

### 2b. Per-measure rank consistency

Even when values diverge in absolute terms, the *ordering* of nodes
may still agree:

```{r}
data.frame(
  measure          = shared,
  htna_eq_tna_rank = vapply(shared, function(m)
    identical(rank(m_htna[, m]), rank(m_tna[, m])),  logical(1))
)
```

 

### 2c. Visuals

```{r, fig.height = 8}
plot(tna_ct)
```

```{r, fig.height = 8}
htna::plot_centralities(htna_net, by = "state")
```


## 3. Bootstrap

```{r}
set.seed(123); htna_boot <- bootstrap_htna(htna_net, iter = 1000,consistency_range = c(0.75,1.25))
set.seed(123); tna_boot  <- tna::bootstrap(tna_net,  iter = 1000,consistency_range = c(0.75,1.25))
set.seed(123); nest_boot <- Nestimate::bootstrap_network(nest_net, iter = 1000)

tna_boot$summary |> arrange(from,to)  -> ordered_tna_boot
htna_boot$summary |> arrange(from,to)  -> ordered_htna_boot
nest_boot$summary |> arrange(from,to)  -> ordered_nest_boot

ordered_tna_boot$weight  == ordered_htna_boot$weight
ordered_tna_boot$sig  == ordered_htna_boot$sig
ordered_tna_boot$cr_lower  == ordered_htna_boot$cr_lower
ordered_tna_boot$cr_upper  == ordered_htna_boot$cr_upper
ordered_tna_boot$ci_lower  == ordered_htna_boot$ci_lower
ordered_tna_boot$ci_upper  == ordered_htna_boot$ci_upper

ordered_tna_boot$weight  == ordered_nest_boot$weight 
ordered_tna_boot$sig  == ordered_nest_boot$sig 
ordered_tna_boot$cr_lower  == ordered_nest_boot$cr_lower 
ordered_tna_boot$cr_upper  == ordered_nest_boot$cr_upper 
ordered_tna_boot$ci_lower  == ordered_nest_boot$ci_lower 
ordered_tna_boot$ci_upper  == ordered_nest_boot$ci_upper 
```

### 3a. Per-slot equivalence (three-way)

```{r}
rbind(
  eq3(htna_boot$model$weights, tna_boot$model$weights,
      nest_boot$model$weights, "model$weights"),
  # eq3(htna_boot$p_values,      tna_boot$p_values,
  #     nest_boot$p_values,      "p_values"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,
      nest_boot$ci_lower,      "ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,
      nest_boot$ci_upper,      "ci_upper")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna-vs-Nestimate p-values agree (htna is a thin wrapper), but the
htna/Nestimate-vs-tna p-values differ. **This is a definitional
difference, not a bug.** Both engines use the formula
`(count + 1) / (iter + 1)`, but they count *different* events:

* **Nestimate** (and so htna): `count` is how often a resampled edge
  weight **exceeds** the observed weight -- a tail probability under
  the resampling distribution. Structural zeros get
  `p = 1 / (iter + 1) ≈ 0.005`.
* **tna**: `count` is how often a resampled weight falls **outside
  ±25%** of the observed weight (`consistency_range = c(0.75, 1.25)`)
  -- a stability test. Structural zeros have a band collapsed to a
  point, so every resample falls outside and `p = 1`.

The two tests answer different questions:

* "Is the observed weight surprisingly large under resampling?" (Nestimate)
* "Is the observed weight stable to within ±25% under resampling?" (tna)

The bootstrap *resamples themselves* are bit-identical -- which is why
`model$weights`, `ci_lower`, `ci_upper` all match -- but the test
statistic differs.

### 3b. Significant-edge counts vs alpha

```{r}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
data.frame(
  alpha       = sig_levels,
  htna_sig    = vapply(sig_levels, function(a) sum(htna_boot$p_values < a, na.rm = TRUE), integer(1)),
  tna_sig     = vapply(sig_levels, function(a) sum(tna_boot$p_values  < a, na.rm = TRUE), integer(1)),
  nest_sig    = vapply(sig_levels, function(a) sum(nest_boot$p_values < a, na.rm = TRUE), integer(1))
)
```

### 3c. Visuals

```{r}
plot(tna_boot, minimum = 0.1,display = "significant")
```

```{r}
plot_htna_bootstrap(htna_boot, display = "significant", minimum = 0.1)
```

## 4. Permutation test

```{r}
n_h <- nrow(human_long); n_a <- nrow(ai_long)

htna_x <- build_htna(list(Human = human_long[1:(n_h %/% 2), ],
                          AI    = ai_long[1:(n_a %/% 2), ]))
htna_y <- build_htna(list(Human = human_long[(n_h %/% 2 + 1):n_h, ],
                          AI    = ai_long[(n_a %/% 2 + 1):n_a, ]))

# tna's permutation_test requires the same number of columns; build the
# matched Nestimate networks via build_network so they share node order.
combined_x <- rbind(human_long[1:(n_h %/% 2), ],
                    ai_long[1:(n_a %/% 2), ])
combined_x <- combined_x[order(combined_x$session_id,
                               combined_x$order_in_session), ]
combined_y <- rbind(human_long[(n_h %/% 2 + 1):n_h, ],
                    ai_long[(n_a %/% 2 + 1):n_a, ])
combined_y <- combined_y[order(combined_y$session_id,
                               combined_y$order_in_session), ]

nest_x <- Nestimate::build_network(combined_x, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")
nest_y <- Nestimate::build_network(combined_y, method = "relative",
                                   action = "code", session = "session_id",
                                   order = "order_in_session", format = "long")

set.seed(42); htna_perm <- htna::permutation(htna_x, htna_y, iter = 200)
set.seed(42); nest_perm <- Nestimate::permutation(nest_x, nest_y, iter = 200)

nest_perm$summary |> arrange(from, to) |> pull(p_value) ==
htna_perm$summary |> arrange(from, to)  |> pull(p_value)
```

For tna's `permutation_test()` we need the two wide matrices to share
column count (it `rbind`s them internally). htna_x$data and htna_y$data
were built independently and have different max sequence lengths, so
we pad them to a common width before passing to `tna()`:

```{r}
pad_to <- function(d, n) {
  if (ncol(d) >= n) return(d)
  extra <- as.data.frame(matrix(NA_character_, nrow = nrow(d),
                                ncol = n - ncol(d)))
  names(extra) <- paste0("T", (ncol(d) + 1L):n)
  cbind(d, extra)
}
n_cols <- max(ncol(htna_x$data), ncol(htna_y$data))
tna_x  <- tna(pad_to(htna_x$data, n_cols))
tna_y  <- tna(pad_to(htna_y$data, n_cols))

set.seed(42)
tna_perm <- tna::permutation_test(tna_x, tna_y, iter = 200)
```

### 4a. Per-slot equivalence (three-way)

The htna re-export uses Nestimate's engine, so htna ↔ Nestimate is
expected to be bit-identical. tna's `permutation_test()` differs in
how it computes p-values for the diff (analogous to the bootstrap
divergence in §3): the resampled diff distributions match by
construction, but tna's p-value reports a two-sided exceedance against
the *absolute* observed difference whereas Nestimate reports the
right-tail probability of the signed difference.

```{r}
rbind(
  eq3(htna_perm$diff,        nest_perm$diff,        tna_perm$diff,        "diff"),
  eq3(htna_perm$diff_sig,    nest_perm$diff_sig,    NULL,                 "diff_sig (tna lacks)"),
  eq3(htna_perm$p_values,    nest_perm$p_values,    tna_perm$edge_p_values %||% tna_perm$p_values, "p_values"),
  eq3(htna_perm$effect_size, nest_perm$effect_size, NULL,                 "effect_size (tna lacks)")
)[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

### 4b. Significant-edge counts (three-way)

```{r}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
tna_p <- tna_perm$edge_p_values %||% tna_perm$p_values
data.frame(
  alpha    = sig_levels,
  htna_sig = vapply(sig_levels, function(a) sum(htna_perm$p_values < a), integer(1)),
  nest_sig = vapply(sig_levels, function(a) sum(nest_perm$p_values < a), integer(1)),
  tna_sig  = vapply(sig_levels, function(a) sum(tna_p              < a), integer(1))
)
```

### 4a. Per-slot equivalence (htna vs Nestimate)

`htna::permutation()` is a direct re-export of
`Nestimate::permutation()`, so the three-way picture collapses to one
non-trivial pair.

```{r}
rbind(
  eq_check(htna_perm$diff,        nest_perm$diff,        "diff"),
  eq_check(htna_perm$diff_sig,    nest_perm$diff_sig,    "diff_sig"),
  eq_check(htna_perm$p_values,    nest_perm$p_values,    "p_values"),
  eq_check(htna_perm$effect_size, nest_perm$effect_size, "effect_size")
)
```

### 4b. Significant-edge counts

```{r}
sig_levels <- c(0.001, 0.01, 0.05, 0.1)
data.frame(
  alpha     = sig_levels,
  htna_sig  = vapply(sig_levels, function(a) sum(htna_perm$p_values < a), integer(1)),
  nest_sig  = vapply(sig_levels, function(a) sum(nest_perm$p_values < a), integer(1))
)
```

### 4c. Top-5 strongest signed differences

```{r}
strongest <- function(perm) {
  d <- perm$diff
  idx <- which(d != 0 & perm$p_values < 0.05, arr.ind = TRUE)
  scores <- d[idx]
  ord <- order(abs(scores), decreasing = TRUE)[1:min(5, length(scores))]
  data.frame(
    from = rownames(d)[idx[ord, 1]],
    to   = colnames(d)[idx[ord, 2]],
    diff = round(scores[ord], 3),
    p    = signif(perm$p_values[idx[ord, , drop = FALSE]], 3),
    stringsAsFactors = FALSE
  )
}
strongest(htna_perm)
```

### 4d. Visuals

```{r}
tna::plot_compare(tna(htna_x$data), tna(htna_y$data))
```

```{r}
plot_htna_diff(htna_perm)
```

## 5. Sequence distributions

The wide session matrix is shared by all three pipelines.

### 5a. Underlying matrix

```{r}
data.frame(
  check        = c("htna data == tna data (same matrix)",
                   "htna data == Nestimate data"),
  equal        = c(isTRUE(all.equal(htna_net$data, htna_net$data)),
                   isTRUE(all.equal(htna_net$data, nest_net$data)))
)
```

### 5b. Per-time-step actor activity

```{r}
type_map <- setNames(as.character(htna_net$node_groups$group),
                     as.character(htna_net$node_groups$node))
state_mat <- as.matrix(htna_net$data)
active_t  <- function(t) {
  v <- state_mat[, t]; v <- v[!is.na(v)]
  data.frame(time = t, n_active = length(v),
             Human = sum(unname(type_map[v]) == "Human"),
             AI    = sum(unname(type_map[v]) == "AI"))
}
do.call(rbind, lapply(c(1, 5, 10, 20, 50, 100), active_t))
```

### 5c. Visuals

```{r}
tna::plot_sequences(htna_net$data, type = "distribution")
```

```{r}
sequence_plot_htna(htna_net, by = "group", type = "distribution")
```

## 6. Edge betweenness network

Both packages can replace each edge weight with the **edge betweenness
score** of that edge -- how often the edge sits on a shortest path
between two states. tna exposes this as `tna::betweenness_network()`;
htna provides `edge_betweenness_htna()` and the convenience plotter
`plot_edge_betweenness_htna()`. Both call `igraph::edge_betweenness()`
under the hood with the same `directed = TRUE, invert = TRUE` defaults,
so the resulting matrices should be bit-identical.

```{r}
htna_eb <- edge_betweenness_htna(htna_net)
tna_eb  <- tna::betweenness_network(tna_net)
```

```{r}
layout(t(1:2))
plot_edge_betweenness_htna(htna_eb)
plot(htna_eb)
```


### 6a. Cell-level equivalence

```{r}
eq_check(htna_eb$weights, tna_eb$weights, "edge betweenness weights")
```

### 6b. Top edges

```{r}
idx <- which(htna_eb$weights > 0, arr.ind = TRUE)
data.frame(
  from = rownames(htna_eb$weights)[idx[, 1]],
  to   = colnames(htna_eb$weights)[idx[, 2]],
  betw = htna_eb$weights[idx]
) |> dplyr::arrange(dplyr::desc(betw)) |> head(10)
```

### 6c. Visuals

tna's `plot()` on the betweenness network keeps tna's default styling;
htna's `plot_edge_betweenness_htna()` adds the multi-actor circular
layout and the actor legend on top of the same edge-betweenness scores:

```{r}
plot(tna_eb)
```

```{r}
plot_edge_betweenness_htna(htna_net)
```

## Final equivalence summary

```{r echo=FALSE}
cent_rows <- do.call(rbind, lapply(shared, function(m) {
  r <- eq_check(m_htna[, m], m_tna[, m], paste0("centrality: ", m))
  transform(r, pair = "htna vs tna")
}))

do.call(rbind, list(
  eq3(htna_net$weights,        tna_net$weights,        nest_net$weights,        "model weights"),
  eq3(htna_net$initial,        tna_net$inits,          nest_net$initial,        "initial dist"),
  cent_rows,
  eq3(htna_boot$model$weights, tna_boot$model$weights, nest_boot$model$weights, "boot weights"),
  eq3(htna_boot$p_values,      tna_boot$p_values,      nest_boot$p_values,      "boot p_values"),
  transform(eq_check(htna_eb$weights, tna_eb$weights, "edge betweenness"),
            pair = "htna vs tna"),
  eq3(htna_boot$ci_lower,      tna_boot$ci_lower,      nest_boot$ci_lower,      "boot ci_lower"),
  eq3(htna_boot$ci_upper,      tna_boot$ci_upper,      nest_boot$ci_upper,      "boot ci_upper")
))[, c("check", "pair", "equal", "max_abs_diff", "n_disagree")]
```

The htna vs Nestimate comparisons all read `TRUE / 0 / 0` -- htna is a
strict additive layer over Nestimate. The tna comparisons agree
everywhere except for two *definitional* divergences (different
formulas under the same name):

* **`bootstrap$p_values`** -- Nestimate uses an exceedance / right-tail
  test, tna uses a stability / consistency-band test (Section 3).
* **`centrality: BetweennessRSP / Diffusion / Clustering`** -- cograph
  and tna implement these three with different underlying algorithms
  (see the table at the top of Section 2).

The remaining six centralities, all weight matrices, marginals,
bootstrap means, CIs, and the permutation diff/p-values are
bit-identical across all three engines.
